---
title: "Project Part 2"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

**Due:** Friday, March 6, 11:59pm on Canvas

**Submission:** A template file is provided, with a rough outline: [project_part_2_template.qmd](https://sta112-s26.github.io/projects/project_part_2_template.qmd)

Once you have completed the report, submit *both* your 
rendered HTML file and the Quarto file to Canvas.

# Overview

Over the course of this semester, you will complete a project in which you use the tools we learn in class to investigate research questions for a given dataset. You will then write a report that communicates the results of your analysis and draws a conclusion about the research questions.

This project will be broken into several different parts; I will provide feedback on each part.

In this second part, your task is to begin to address the research questions by fitting an appropriate linear regression model.

# Data

The [College Scorecard](https://collegescorecard.ed.gov/), compiled by the US Department of Education, is meant to be a tool for college applicants to compare different institutions. In this project, you will work with data from the scorecard to advise a high school senior on choosing a college. (We will use data that are several years old, so may be slightly out of date, but we will hope that the same general trends are still true today).

## Research questions

For this project, suppose you have been approached by a high school senior applying to college. They are looking through the college scorecard data, and they have some questions for you about the value of different schools. 

1. Do students who graduate from more expensive schools earn more money after graduation?

2. Is the relationship between school cost and graduate earnings the same for public and private schools?

You will investigate each of these research questions in your analysis, and explain your results in your report.

## Variables

The dataset provided for this project contains the following columns:

* `INSTNM`: school name
* `STABBR`: state in which the school is located
* `CONTROL`: whether the school is public or private
* `SATVRMID`: midpoint of SAT critical reading scores of students attending the school
* `SATMTMID`: midpoint of SAT math scores
* `SATWRMID`: midpoint of SAT writing scores
* `ACTCMMID`: midpoint of the ACT cumulative scores
* `ACTENMID`: midpoint of ACT English scores
* `ACTMTMID`: midpoint of ACT math scores
* `ACTWRMID`: midpoint of ACT writing scores
* `UGDS`: number of undergraduate students at the school
* `NPT4`: average cost to attend the school
* `PCTFLOAN`: fraction of undergraduates receiving a federal student loan
* `MD_EARN_WNE_P10`: median salary of students 10 years after graduation
* `GRAD_DEBT_MDN_SUPP`: median debt of graduated students
* `RPY_3YR_RT_SUPP`: proportion of students who are actively repaying their loans 3 years after graduation
* `RPY_3YR_70`: whether the proportion of students actively repaying their loans 3 years after graduation is > 70% (1 = yes, 0 = no)

## Downloading the data

The data for this project is stored in a CSV file, and can be imported into R with the `read_csv` function in the `readr` package. The following code imports the data and saves it as a data frame called `scorecard`:

```r
scorecard <- read_csv("https://sta112-s26.github.io/projects/scorecard.csv")
```

# Your task in Part II

Based on your exploratory data analysis from Part 1, fit a linear regression model to address **the first research question**: *Do students who graduate from more expensive schools earn more money after graduation?*. Then, use a residual plot to assess the shape and constant variance assumptions, and a studentized residual plot to check for potential outliers. Make any additional adjustments to your model based on your diagnostic plots. Interpret the fitted model: discuss the estimated slope in context of the research question, and report and interpret the model's RMSE. Finally, perform inference (a hypothesis test and confidence interval) to address the research question of interest.

There is no minimum or maximum length required. However, your analysis should be thorough, with your report addressing the research question, and containing all the elements described below.

## Model

First, we need to describe the regression model and give the equation of the fitted line. Write a short paragraph in which you:

* Give the population regression model you will use to address the research question. This should be written using correct notation (remember that $\varepsilon$ is included in the population model, and there are no hats on any variables or parameters). Make sure to use informative variable names. See "Writing math in Quarto" below.
* Give the equation of the fitted regression line, using correct notation and informative variable names (remember to provide the actual values of $\widehat{\beta}_0$ and $\widehat{\beta}_1$).

## Assumption Diagnostics

Next, we want to assess whether the regression assumptions are reasonable for our model. Write 1 - 2 paragraphs in which you:

* Create a residual plot and studentized residual plot for your fitted model. Make sure to include informative axis labels and captions (with figure numbers) for your figures.
* Use your residual plot to discuss whether the shape and constant variance assumptions are reasonable. Refer to your figures by their numbers in the text.
* Use your studentized residual plot to identify any potential outliers
* Make any changes to your regression model based on the residual plot; if you have made changes, report the equation of your final fitted model

## Interpretation

Write a short paragraph in which you:

* Interpret the estimated slope in your final fitted model, in context of the data and research question
* Discuss the RMSE of the fitted model

## Inference

Finally, we want to use our fitted regression model to address the research question. Write 1-2 paragraphs in which you:

* Translate the research question into null and alternative hypotheses about one or more model parameters. Write these hypotheses using correct notation.
* Calculate a test statistics and p-value for your hypotheses.
* Use the p-value to make a conclusion about the research question
* Report and interpret a confidence interval for that parameter of interest

## Sources

Provide information on any sources you have used outside the course materials.

* Cite any articles you have referenced
* Provide links to any websites you have used for code help
* If generative AI (Chat GPT, Gemini, etc.) was used to help with any part of the assignment, summarize
    * The AI tool used
    * The prompts you used
    * The help provided by the tool


# Appearance and style

### Creating a studentized residual plot

Here is example code for creating a studentized residual plot, using the Florida voting data as an example:

```{r, message=F, warning=F}
library(tidyverse)
library(Stat2Data)
data("PalmBeach")

vote_lm <- lm(Buchanan ~ Bush, data = PalmBeach)

data.frame(studentized_resids = rstudent(vote_lm),
         yhat = predict(vote_lm)) |>
  ggplot(aes(x = yhat, y = studentized_resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd=1.2) +
  geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
              color = "orange", lty = 2, lwd=1.2) +
  labs(x = "Predicted votes for Buchanan",
       y = "Studentized residuals") +
  theme_bw()
```


### Writing

The report should be written like an article or research paper: in full sentences and paragraphs, with headings for each section. You should not write your report with question numbers or as a list of bullet points. Scientific articles are generally written in third person, though "we" can also be acceptable ("we can see from Figure 1...") in some disciplines.

### Code

As with previous assignments, you will write the document in Quarto. However, in full reports, the only output that should be visible from code chunks are figures and tables. If a code chunk does not produce a figure or table, you can hide it from the knitted document with `include=F`:

````
```{r, include=F}`r ''`

```
````

If a code chunk produces a figure or table, only the figure or table should be visible in the knitted document. You can hide the chunk but display the output with `echo=F, message=F, warning=F`:

````
```{r, echo=F, message=F, warning=F}`r ''`

```
````

### Figures

Figures should have labeled axes, and should be clear and easy to read. Figures should also be captioned and numbered; to caption a figure, use `fig.cap = "..."` in the chunk options. For example (scroll to the right on the code to see it all),

````
```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}`r ''`
penguins |>
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```
````

is displayed as

```{r include=F}
library(tidyverse)
library(palmerpenguins)
```

```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}
penguins |>
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```

Captions should provide enough information to understand what is being plotted, but interpretation can be left to the main text. Refer to figures by their number in the text. Make sure that any figures you include are discussed in the text.

### Writing math in Quarto documents

Doing things like writing down models often involve parameters. Parameters in statistics are generally represented as Greek letters. How can we write Greek letters and other mathematical notation in our Markdown file so they show up when we knit?

If you want to write mathematical notation, we need to tell Markdown, “Hey, we’re going to make a math symbol!” To do that, you use dollar signs. For instance, to make $\widehat{\beta}_1$, you simply put `$\widehathat{\beta}_1$` into the white space (not a chunk) in your Markdown.

Here are some examples of writing math, which you can adapt:

| Math | Code |
| --- | --- |
| $Y = \beta_0 + \beta_1 X + \varepsilon$ | `$Y = \beta_0 + \beta_1 X + \varepsilon$` |
| $\widehat{Y} = -2.2 + 3.7 X$ | `$\widehat{Y} = -2.2 + 3.7 X$` |

# Checklist

* Change name in Quarto YAML
* Five sections: Model, Diagnostics, Interpretation, Inference, and Sources (see descriptions above for requirements)
* Writing:
  * Each section written as a paragraph, with full sentences
  * No bullet points or lists
  * Math equations are written nicely, with correct notation (see "Writing math in Quarto")
* Code:
  * Code is included in the Quarto document, but *not* displayed in the rendered HTML
  * The only R output shown in the report is figures and tables
* Figures:
  * Include meaningful caption and axis labels (see above for details on including a caption)
  * Each figure included in the report is discussed in your writing, and referred to by its figure number
* Submission:
  * Submit *both* the rendered HMTL and the .qmd file to Canvas
  * Deadline is Friday, March 6 at 11:59pm