---
title: Interactions
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(palmerpenguins)
```


### Last time: Class activity

.large[
\begin{align}
\widehat{\text{Height}} &= 40.689 - 7.728 \ \text{IsMale} + 0.118 \ \text{Age} \\
& \hspace{1cm} + 0.064 \ \text{Age} \cdot \text{IsMale}
\end{align}

.question[
What is the estimated height for a girl aged 120 months (10 years)?
]
]

---

### Last time: Class activity

.large[
\begin{align}
\widehat{\text{Height}} &= 40.689 - 7.728 \ \text{IsMale} + 0.118 \ \text{Age} \\
& \hspace{1cm} + 0.064 \ \text{Age} \cdot \text{IsMale}
\end{align}

.question[
What is the estimated height for a girl aged 120 months (10 years)?
]
]

.large[
$40.689 + 0.118(120) = 54.85$ inches (about 4 feet 7 inches)
]

---

### Last time: Class activity

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6.5, fig.height=4}
library(Stat2Data)
data("Kids198")
kids <- Kids198 %>%
  mutate(Sex = ifelse(Sex == 0, "male", "female"))

kids %>%
  ggplot(aes(x = Age, y = Height, color = Sex)) +
  geom_point() +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
With interaction: $R^2_{adj} = 0.763$

Without interaction: $R^2_{adj} = 0.732$
]

.large[
.question[
Do you think an interaction is needed?
]
]

---

### Last time: Class activity

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7.5, fig.height=4.5}
library(Stat2Data)
data("Kids198")
kids <- Kids198 %>%
  mutate(Sex = ifelse(Sex == 0, "male", "female"))

kids %>%
  ggplot(aes(x = Age, y = Height, color = Sex)) +
  geom_point() +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
I think an interaction is important
* slopes look different 
* captures that boys and girls start off with similar heights, but boys eventually become taller.
]

---

### Interactions with a categorical predictor

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

.large[
The slope on body mass depends on species:

* Slope for Adelie penguins: $\beta_3$
* Slope for Chinstrap penguins: $\beta_3 + \beta_4$
* Slope for Gentoo penguins: $\beta_3 + \beta_5$
]

---

### Interactions with a categorical predictor

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

.large[
The slope on body mass depends on species:

$\text{slope} = \beta_3 + \beta_4 \ \text{IsChinstrap} + \beta_5 \ \text{IsGentoo}$

.question[
We can do something similar for two quantitative variables
]
]

---

### Interactions with two quantitative variables

.large[

Additive model (no interactions):

$$\text{bill length} = \beta_0 + \beta_1 \text{body mass} + \beta_2 \text{flipper length} + \varepsilon$$

* Slope on body mass is $\beta_1$
* Slope on body mass *does not depend on* flipper length
]

---

### Interactions with two quantitative variables

.large[

$$\text{bill length} = \beta_0 + (\text{slope}) \ \text{body mass} + \beta_2 \text{flipper length} + \varepsilon$$

**Making the slope depend on flipper length:**

slope on body mass = $\beta_1 + \beta_3 \text{flipper length}$

**Substituting:**

\begin{align}
\text{bill length} &= \beta_0 + (\beta_1 + \beta_3 \text{flipper length}) \ \text{body mass} \\
& \hspace{1cm} + \beta_2 \text{flipper length} + \varepsilon \\
&= \beta_0 + \beta_1 \text{body mass} + \beta_2 \text{flipper length} \\
& \hspace{1cm} + \beta_3 \text{body mass} \cdot \text{flipper length} + \varepsilon
\end{align}
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{body mass} + \beta_2 \text{flipper length} \\
& \hspace{1cm} + \beta_3 \text{body mass} \cdot \text{flipper length} + \varepsilon
\end{align}
]

.large[
* $\text{body mass} \cdot \text{flipper length}$ is the *interaction* between body mass and flipper length
* As before, it literally means multiplication
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
What is the predicted bill length for a penguin with body mass 5000g and flipper length 200mm?
]
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
What is the predicted bill length for a penguin with body mass 5000g and flipper length 200mm?
]
]

.large[
$-26.75 + 0.0059(5000) + 0.339(200) - 0.000026(5000)(200)$

$= 44.55$
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
What is the change in bill length associated with a 1g increase in body mass, for penguins with flipper length = 200mm?
]
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
What is the change in bill length associated with a 1g increase in body mass, for penguins with flipper length = 200mm?
]
]

.large[
$0.0059 - 0.000026(200) = 0.0007$
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
Is the slope on body mass higher or lower for penguins with longer flippers?
]
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
Is the slope on body mass higher or lower for penguins with longer flippers?
]
]

.large[
Lower, because the coefficient -0.000026 is negative.
]

---

### Visualizing interactions

.large[
With a categorical variable, we can fit a different slope for each group:
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.height=5, fig.width=7}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
.question[
How would this work with two quantitative variables?
]
]

---

### Visualizing interactions

.large[
**Binarize:**

$$LongFlipper = \begin{cases} 1 & \text{flipper length} > 200 \\ 0 & \text{flipper length} \leq 200 \end{cases}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.height=5, fig.width=7}
penguins %>%
  drop_na() %>%
  mutate(LongFlipper = ifelse(flipper_length_mm > 200, "1", "0")) %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = LongFlipper)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)")
```

---

### Homework work time


.large[
Homework 5 involves a research question with an interaction between two quantitative variables. Spend the remainder of the class period getting started on HW 5.
]

