---
title: Introduction to multiple linear regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Some statistics classes next semester

.large[
* STA 214 (Applied Generalized Linear Models)
  * Requires MTH 111 (Calc 1)
* STA 247 (Design and Sampling)
* STA 261 (Statistical Computing)
* STA 279 (Text Data)
* STA 310 (Probability)
  * Requires MTH 112 (Calc 2)
* Also consider MTH 121 (Linear algebra) for upper-level stats courses
]

---

### Last time

.large[
\begin{align}
\log(\text{Price per carat}) &= \beta_0 + \beta_1 \ \text{IsSI1} + \beta_2 \ \text{IsSI2} + \beta_3 \ \text{IsSI3} \ + \\ & \hspace{1cm} \beta_4 \ \text{IsVS1} + \beta_5 \ \text{IsVS2} + \beta_6 \ \text{IsVVS1} + \\ & \hspace{1cm} \beta_7 \ \text{IsVVS2} + \varepsilon
\end{align}

.question[
 We want to test whether there is a relationship between Clarity and price per carat. What are our null and alternative hypotheses?
]
]

---

### Last time

.large[
$H_0: \beta_1 = \beta_2 = \cdots = \beta_7 = 0$
$H_A: \text{at least one of } \beta_1, \beta_2, ..., \beta_7 \neq 0$

```
           Df Sum Sq Mean Sq F value  Pr(>F)  
Clarity     7  3.038 0.43400  2.3832 0.02164 *
Residuals 343 62.463 0.18211                  
```
]

---

### Last time

.large[
```
            Estimate Std. Error t value Pr(>|t|)    
(Intercept)  1.81518    0.10669  17.014   <2e-16 ***
ClaritySI1  -0.21875    0.12579  -1.739   0.0829 .  
ClaritySI2  -0.24783    0.16714  -1.483   0.1391    
ClaritySI3  -0.82695    0.32006  -2.584   0.0102 *  
ClarityVS1  -0.01472    0.11347  -0.130   0.8969    
ClarityVS2  -0.07700    0.11550  -0.667   0.5054    
ClarityVVS1 -0.02433    0.13136  -0.185   0.8532    
ClarityVVS2 -0.13451    0.12878  -1.044   0.2970    
```

.question[
Suppose that instead I just want to know whether there is a difference in price per carat between Internally Flawless (IF) diamonds and Very Slightly Included (VS1) diamonds. What would my hypotheses be for this test? How would I perform the test?
]
]

---

### Relationships between > 2 variables

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=5}
library(palmerpenguins)
library(tidyverse)
library(latex2exp)
library(Stat2Data)
library(knitr)

penguins %>%
  drop_na() %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

.large[
.question[
How can we model the relationship between body mass, species, and bill length?
]
]

---

### Regression with a single quantitative predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

penguins %>%
  drop_na() %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm)) +
  geom_point() +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
**Model:**

.center[
$\text{bill length} = \beta_0 + \beta_1 \text{body mass} + \varepsilon$
]
]

---

### Regression with a single categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

penguins %>%
  drop_na() %>%
  ggplot(aes(x = species,
             y = bill_length_mm)) +
  geom_boxplot() +
  labs(x = "Species",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```


.large[
**Model:**

.center[
$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$
]
]

---

### Regression with $> 2$ variables

.large[
$\text{bill length} = \beta_0 + \beta_1 \text{body mass} + \varepsilon$

$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$

.question[
How can we include both species *and* body mass in the model?
]
]

---

## Multiple regression

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}

penguins <- penguins |>
  drop_na(bill_length_mm, species, body_mass_g)

length_lm <- lm(bill_length_mm ~ species + body_mass_g, 
                data = penguins)

penguins %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_line(aes(y = yhat), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```


.large[
**Model:**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}
]

---

### Multiple regression model

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**This gives us a separate regression line for each species of penguin.**

.question[
What is the regression line for Adelie penguins?
]
]

---

### Multiple regression model

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**This gives us a separate regression line for each species of penguin.**

.question[
What is the regression line for Adelie penguins?
]
]

.large[
$\text{bill length} = \beta_0 + \beta_3 \text{body mass} + \varepsilon$

.question[
What about for Chinstrap and Gentoo penguins?
]
]

---

### Multiple regression model

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**This gives us a separate regression line for each species of penguin:**

$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \beta_3 \text{body mass} + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Gentoo} \end{cases}$$

]

---

### Fitted model

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.92 + 9.92 \ \text{IsChinstrap} + 3.56 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0037 \ \text{body mass}
\end{align}

.question[
How would I interpret $\widehat{\beta}_0 = 24.92$?
]
]

---

### Fitted model

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.92 + 9.92 \ \text{IsChinstrap} + 3.56 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0037 \ \text{body mass}
\end{align}

.question[
How would I interpret $\widehat{\beta}_0 = 24.92$?
]
]

.large[
The estimated bill length for an Adelie penguin weighing 0g. (not that meaningful)

.question[
What about $\widehat{\beta}_1 = 9.92$ and $\widehat{\beta}_2 = 3.56$?
]
]

---

### Interpreting regression coefficients

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
library(latex2exp)
penguins %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm)) +
  geom_point(size = 2, alpha = 0) +
  geom_line(aes(y = yhat, color = species), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  geom_segment(aes(x = 3500, y = 38,
                   xend = 3500, yend = 48),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 4500, y = 42,
                   xend = 4500, yend = 45),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 3000, y = 41, 
           label = TeX("$9.92", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 5000, y = 44, 
           label = TeX("$3.56", output='character'), 
           parse=TRUE, size = 7) +
  theme_bw() +
  theme(text = element_text(size = 25))
```


.large[

* $9.92 =$ estimated mean difference in bill length for a Chinstrap and Adelie penguin with the same body mass
* $3.56 =$ estimated mean difference in bill length for a Gentoo and Adelie penguin with the same body mass
]

---

### Interpreting regression coefficients

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.92 + 9.92 \ \text{IsChinstrap} + 3.56 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0037 \ \text{body mass}
\end{align}

.question[
How would I interpret $\widehat{\beta}_3 = 0.0037$?
]
]

---

### Interpreting regression coefficients

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.92 + 9.92 \ \text{IsChinstrap} + 3.56 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0037 \ \text{body mass}
\end{align}

.question[
How would I interpret $\widehat{\beta}_3 = 0.0037$?
]
]

.large[
The average change in bill length associated with an increase of 1g in body mass, holding species constant.

*Note that we assume the slope is the same for each species!*
]

---

### Predictions

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.92 + 9.92 \ \text{IsChinstrap} + 3.56 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0037 \ \text{body mass}
\end{align}

.question[
What is the predicted bill length for a Gentoo penguin with a body mass of 5000 g?
]
]

---

### Predictions

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.92 + 9.92 \ \text{IsChinstrap} + 3.56 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0037 \ \text{body mass}
\end{align}

.question[
What is the predicted bill length for a Gentoo penguin with a body mass of 5000 g?
]
]

.large[
$24.92 + 3.56 + 0.0037(5000) = 46.98$
]

---

### Fitting the model in R

.large[
```{r}
m1 <- lm(bill_length_mm ~ species + body_mass_g, 
                 data = penguins)
```

* `bill_length_mm ~ species + body_mass_g` is a `formula` in R
* Still `response ~ predictor(s)`
* `+` means "include more variables in the model"
]

---

### Fitting the model in R

.large[
```{r, output.lines = 10:14}
m1 <- lm(bill_length_mm ~ species + body_mass_g, 
                 data = penguins)
summary(m1)
```

\begin{align}
\widehat{\text{bill length}} &= 24.92 + 9.92 \ \text{IsChinstrap} + 3.56 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0037 \ \text{body mass}
\end{align}
]

---

### Class activity

.large[
Work on the class activity (handout).
]

---

### Class activity

.large[
.question[
What is the population regression model?
]
]

---

### Class activity

.large[
.question[
What is the population regression model?
]
]

.large[
\begin{align}
\text{Weight} &= \beta_0 + \beta_1 \text{IsEnlarged} + \beta_2 \text{IsReduced} + \\
& \hspace{1cm} \beta_3 \text{WingLength} + \varepsilon
\end{align}
]

---

### Class activity

.large[

\begin{align}
\widehat{\text{Weight}} &= 1.94 - 0.46 \text{IsEnlarged} + 0.90 \text{IsReduced} + \\
& \hspace{1cm} 0.45 \text{WingLength}
\end{align}

.question[
What is the average change in weight associated with a 1mm increase in wing length, for sparrows from enlarged nests?
]
]

---

### Class activity

.large[

\begin{align}
\widehat{\text{Weight}} &= 1.94 - 0.46 \text{IsEnlarged} + 0.90 \text{IsReduced} + \\
& \hspace{1cm} 0.45 \text{WingLength}
\end{align}

.question[
What is the average change in weight associated with a 1mm increase in wing length, for sparrows from enlarged nests?
]
]

.large[
Answer: 0.45

(We fit the same slope for each group of sparrows, and it is estimated to be 0.45)
]

---

### Class activity

.large[

\begin{align}
\widehat{\text{Weight}} &= 1.94 - 0.46 \text{IsEnlarged} + 0.90 \text{IsReduced} + \\
& \hspace{1cm} 0.45 \text{WingLength}
\end{align}

.question[
What is the estimated difference in weight for sparrows from enlarged vs. control nests, with the same wing length?
]
]

---

### Class activity

.large[

\begin{align}
\widehat{\text{Weight}} &= 1.94 - 0.46 \text{IsEnlarged} + 0.90 \text{IsReduced} + \\
& \hspace{1cm} 0.45 \text{WingLength}
\end{align}

.question[
What is the estimated difference in weight for sparrows from enlarged vs. control nests, with the same wing length?
]
]

.large[
We estimate that sparrows from enlarged nests weigh 0.46g less, on average, than sparrows from control nests with the same wing length
]

