---
title: Regression and hypothesis testing with a categorical explanatory variable
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(palmerpenguins)
library(tidyverse)
library(latex2exp)
library(Stat2Data)

penguins_no_nas <- penguins %>%
  drop_na()

penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1.5) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20))
```

.large[
* How do we model the relationship between species and bill length?
* Can we test for a relationship between bill length and species?
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$$

.question[
How can I write this more concisely?
]
]

---

### Indicator variables

.large[
$$\text{IsChinstrap} = \begin{cases} 0 & \text{species } \neq \text{Chinstrap} \\ 1 & \text{species = Chinstrap} \end{cases}$$

$$\text{IsGentoo} = \begin{cases} 0 & \text{species } \neq \text{Gentoo} \\ 1 & \text{species = Gentoo} \end{cases}$$

Then:

| Species | IsChinstrap | IsGentoo |
| --- | --- | --- |
| Adelie | 0 | 0 |
| Chinstrap | 1 | 0 |
| Gentoo | 0 | 1 |
]

---

### Indicator variables

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=3.5}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$$

Or, more concisely:

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$
]

---

### Model assumptions

.large[

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

.question[
What model assumptions do you think we make?
]
]

---

### Model assumptions

.large[

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

where $\varepsilon \sim N(0, \sigma_{\varepsilon}^2)$ and the $\varepsilon$ are independent of each other.

* No notion of shape assumption with single categorical predictor
* Still assume constant variance, normality, independence, and randomness

.question[
How should we assess each of these assumptions?
]
]

---

### Assessing constant variance

.large[
For a single categorical predictor, we can create a boxplot of residuals for each group, and calculate the standard deviation of residuals in each group.
]

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguin_lm <- lm(bill_length_mm ~ species, 
                 data = penguins_no_nas)

penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(x = species, y = resids)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Species", 
       y = "Residuals") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
```{r echo=F, message=F, warning=F}
library(knitr)
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  group_by(species) %>%
  summarize(residual_std_dev = round(sd(resids), 2)) %>%
  kable()
```
]
]

---

### Assessing constant variance

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguin_lm <- lm(bill_length_mm ~ species, 
                 data = penguins_no_nas)

penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(x = species, y = resids)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Species", 
       y = "Residuals") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
```{r echo=F, message=F, warning=F}
library(knitr)
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  group_by(species) %>%
  summarize(residual_std_dev = round(sd(resids), 2)) %>%
  kable()
```
]
]

.large[
Looking for similar variability between groups. Rule of thumb: not too concerned about constant variance if

$$\dfrac{\text{largest standard deviation}}{\text{smallest standard deviation}} < 2$$
]

---

### Assessing normality

.large[
Can use a QQ plot again:
]

```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq() +
  geom_qq_line() +
  theme_bw() +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme(text = element_text(size = 20))
```

---

### Statistical inference

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1.5) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 20))
```

.large[
.question[
Does it look like there is a relationship between species and bill length?
]
]

---

### Hypothesis testing

.large[
$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

We want to know whether there is a relationship between species and bill length.

.question[
Can I translate this into null and alternative hypotheses about one or more model parameters?
]
]

---

### Hypothesis testing

.large[
$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

We want to know whether there is a relationship between species and bill length.

.question[
Can I translate this into null and alternative hypotheses about one or more model parameters?
]
]

.large[
$H_0: \beta_1 = \beta_2 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$
]

---

### Full and reduced models

.large[
$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$
$H_0: \beta_1 = \beta_2 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$


**Reduced model ( $H_0$ ):**

$$\text{bill length} = \beta_0 + \varepsilon$$

**Full model ( $H_A$ ):**

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

.question[
What do I need to test these hypotheses?
]
]

---

### Intuition

.large[
.question[
Which of these two plots would show more evidence of a relationship between species and bill length?
]
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(bill_length_mm = bill_length_mm + 
           ifelse(species == "Adelie", 0, 
                  ifelse(species == "Chinstrap", 10, 5))) %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25),
        axis.text.y = element_blank())
```

]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(bill_length_mm = bill_length_mm + 
           rnorm(nrow(penguins_no_nas), sd=20)) %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25),
        axis.text.y = element_blank())
```

]

---

### Intuition

.large[
More evidence for a relationship when *between-group* variability is larger than *within-group* variability.
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(bill_length_mm = bill_length_mm + 
           ifelse(species == "Adelie", 0, 
                  ifelse(species == "Chinstrap", 10, 5))) %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25),
        axis.text.y = element_blank())
```

.large[
Large between-group variability, relative to within-group variability.
]
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(bill_length_mm = bill_length_mm + 
           rnorm(nrow(penguins_no_nas), sd=20)) %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25),
        axis.text.y = element_blank())
```

.large[
Small between-group variability, relative to within-group variability.
]
]

---

### Between-group vs. within-group variability

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
**Between-group variability:** $\sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2$ 

**Within-group variability:** $\sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2$ 
]

---

### Partitioning variability and degrees of freedom

.large[
**Variability:** $\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel \hspace{0.5cm} + \hspace{0.5cm} SSE$

**Degrees of freedom:** 

$\hspace{4cm} n-1 \hspace{0.5cm} = \hspace{0.5cm} (p - 1)  \hspace{0.5cm} + \hspace{0.5cm} (n - p)$
]

---

### Between-group vs. within-group variability

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
**Test statistic:** $F = \dfrac{\text{between group variance}}{\text{within group variance}} = \dfrac{SSModel/(p-1)}{SSE/(n-p)}$
]

---

### Partitioning variability

.large[
**Variability:** $\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel \hspace{0.5cm} + \hspace{0.5cm} SSE$


**Test statistic:** $F = \dfrac{\frac{1}{p-1} \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2}{\frac{1}{n-p} \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2} = \dfrac{\frac{1}{p-1} SSModel}{\frac{1}{n-p} SSE} = \dfrac{MSModel}{MSE}$ 
]

---

### Analysis of variance

.large[

**Test statistic:** $F = \dfrac{\frac{1}{p-1} \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2}{\frac{1}{n-p} \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2} = \dfrac{\frac{1}{p-1} SSModel}{\frac{1}{n-p} SSE} = \dfrac{MSModel}{MSE}$ 

**Analysis of variance (ANOVA) table:**

| Source | df | SS | MS | F |
| --- | --- | --- | --- | --- |
| Model | $p-1$ | $SSModel$ | $MSModel = \frac{SSModel}{p-1}$ | $F = \frac{MSModel}{MSE}$ |
| Residual | $n - p$ | $SSE$ | $MSE = \frac{SSE}{n-p}$ | |
| Total | $n-1$ | $SSTotal$ | | |
]

---

### Example

.large[

$$\sum \limits_{i=1}^n (\widehat{\text{bill length}}_i - \overline{\text{bill length}})^2 = 7015.4 \hspace{1cm}$$ 

$$\sum \limits_{i=1}^n (\text{bill length}_i - \widehat{\text{bill length}}_i)^2 = 2913.5 \hspace{1cm}$$

| Source | df | SS | MS | F |
| --- | --- | --- | --- | --- |
| Model | $\hspace{2cm}$ | $\hspace{4cm}$  | $\hspace{4cm}$ | $\hspace{4cm}$ |
| Residual | |  |  | |
| Total |  | | | |
]

---

### Class activity

.large[
Work on the class activity (handout).
]



