---
title: Regression with a categorical predictor
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Previously: regression with a categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=3.5}
library(tidyverse)
library(palmerpenguins)
library(latex2exp)

set.seed(1)
penguins %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$$

Or, more concisely:

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$
]

---

### Previously: regression with a categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=3.5}
penguins %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1.5) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
.question[
Can I use the correlation to summarize the relationship between species and bill length?
]
]

---

### Previously: polynomial transformations

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
library(Stat2Data)
data("CO2Germany")
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$$\widehat{\text{CO2}} = 414.975 - 0.476 \ \text{Day} + 0.0012 \ \text{Day}^2$$
]

---

### Previously: polynomial transformations

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
library(Stat2Data)
data("CO2Germany")
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$$\widehat{\text{CO2}} = 414.975 - 0.476 \ \text{Day} + 0.0012 \ \text{Day}^2$$

.question[
Can I use the correlation to calculate the relationship between Day and CO2?
]
]

---

### Measuring variability in the response

.large[
* When we fit a regression model with least squares, we choose parameters to minimize the sum of squared residuals:

$$SSE = \sum_i (y_i - \widehat{y}_i)^2$$
]

.large[
* We can think of SSE as the "amount of variability" in $Y$ after predicting with our regression model
* The amount of variability in $Y$ *before* predicting with our regression model is

$$SSTotal = \sum_i (y_i -\overline{y})^2$$
]

---

### Thinking about SSE

.large[
$$SSE = \sum_i (y_i - \widehat{y}_i)^2$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

---

### Thinking about SSTotal

.large[
$$SSTotal = \sum_i (y_i -\overline{y})^2$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ 1, lwd=1.2)
```

---

### SSE vs. SSTotal

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)

m1 <- lm(CO2 ~ poly(Day, 2, raw=TRUE), data = CO2Germany)
#sum(m1$residuals^2)
```

SSE = 4992.71
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ 1, lwd=1.2)

m2 <- lm(CO2 ~ 1, data = CO2Germany)
#sum(m2$residuals^2)
```

SSTotal: 11703.69
]

.large[
.question[
Do I want SSE to be large relative to SSTotal, or small relative to SSTotal?
]
]

---

### $R^2$: the coefficient of determination

.large[
$$SSE = \sum_i (y_i - \widehat{y}_i)^2$$
$$SSTotal = \sum_i (y_i -\overline{y})^2$$
]

.large[
The proportion of variability in $Y$ "explained" by our regression model is 

$$R^2 = 1 - \dfrac{SSE}{SSTotal}$$
]

.large[
* $R^2$ is between 0 and 1
* A higher $R^2$ means our model is a "better fit" to the data
* For regression with one variable, $R^2 = r^2$ (the squared correlation)
]

---


### SSE vs. SSTotal

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)

m1 <- lm(CO2 ~ poly(Day, 2, raw=TRUE), data = CO2Germany)
#sum(m1$residuals^2)
```

SSE = 4992.71
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ 1, lwd=1.2)

m2 <- lm(CO2 ~ 1, data = CO2Germany)
#sum(m2$residuals^2)
```

SSTotal: 11703.69
]

.large[
$R^2 = 1 - \dfrac{4992.71}{11703.69} = 0.573$
]

---

### SSE vs. SSTotal

.pull-left[
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
penguins %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0.5, size=2) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38.8, xend = 1.4, yend = 38.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 48.8, xend = 2.4, yend = 48.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47.6, xend = 3.4, yend = 47.6), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

SSE = 2969.9
]

.pull-right[
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
penguins %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0.5, size=2) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 44, xend = 3.4, yend = 44), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

SSTotal = 10164.21
]

.large[
$R^2 = 1 - \dfrac{2969.9}{10164.21} = 0.708$
]

---

### Comparing $R^2$ values

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$R^2 = 0.573$
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
$R^2 = 0.575$
]
]

.large[
.question[
$R^2$ has increased very slightly -- does that mean that a polynomial of order 3 is a better fit?
]
]

---

### $R^2$

.large[
$R^2$ **will always increase when you add more parameters to the model.**

Why? Because

$$R^2 = 1 - \dfrac{SSE}{SSTotal}$$
and adding more parameters gives us more flexibility to decrease SSE.
]

---

### Adjusted $R^2$

.large[
We can account for the number of parameters in the model with *adjusted* $R^2$:

$$R^2_{adj} = 1 - \dfrac{SSE/(n-p)}{SSTotal/(n-1)}$$
]

.large[
* $n =$ number of observations in data (sample size)
* $p =$ number of parameters ( $\beta$ terms) in model
]

---

### Adjusted $R^2$

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
SSE = 4992.7

SSTotal = 11703.69
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
SSE = 4971.9

SSTotal = 11703.69
]
]

.large[
.question[
Why is SSTotal the same for the two models?
]
]

---

### Adjusted $R^2$

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
SSE = 4992.7

SSTotal = 11703.69
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
SSE = 4971.9

SSTotal = 11703.69
]
]

.large[
$$R^2_{adj} = 1 - \dfrac{SSE/(n-p)}{SSTotal/(n-1)}$$
]

---

### Adjusted $R^2$

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$R^2_{adj} = 0.56976$
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
$R^2 = 0.56972$
]
]

---

### Properties of $R^2_{adj}$

.large[
* $R^2_{adj} < R^2$ (always)
* $R^2_{adj}$ accounts for the number of parameters in the model
* $R^2$ can *never* decrease when predictors are added
* $R^2_{adj}$ *can* decrease if unhelpful predictors are added
]

---

### Values in R

.large[
```{r, eval=F}
penguin_mod <- lm(bill_length_mm ~ species, 
                  data = penguins)
summary(penguin_mod)
```

```
                 Estimate Std. Error t value Pr(>|t|)    
(Intercept)       38.7914     0.2409  161.05   <2e-16 ***
speciesChinstrap  10.0424     0.4323   23.23   <2e-16 ***
speciesGentoo      8.7135     0.3595   24.24   <2e-16 ***
...
Multiple R-squared:  0.7078, Adjusted R-squared:  0.7061
```

```{r, eval=F}
co2_mod <- lm(CO2 ~ poly(Day, 2, raw=TRUE), 
              data = CO2Germany)
summary(co2_mod)
```

```
...
Multiple R-squared:  0.5734, Adjusted R-squared:  0.5698
```
]



---

### Class activity

.large[
* Practice regression with a single categorical predictor, with diamonds data
* Calculate $R^2$ and $R^2_{adj}$
* No need to submit this one on Canvas

[https://sta112-s26.github.io/class_activities/ca_16.html](https://sta112-s26.github.io/class_activities/ca_16.html)
]



