---
title: Outliers and influential points
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---


### Anything unusual?

.large[
Votes for Bush and Buchanan, in Florida counties in 2000:
]

```{r echo=F, message=F, warning=F, fig.width = 7, fig.height=5, fig.align='center', message=F}
library(tidyverse)
library(Stat2Data)
data("PalmBeach")

PalmBeach %>%
  ggplot(aes(x = Bush, y = Buchanan)) +
  geom_point(size = 2.5) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
.question[
Does anything look unusual in this scatterplot?
]
]

---

### Anything unusual?

.large[
**Outlier:** A point which doesn't follow the general relationship between the predictor and response.
]

```{r echo=F, message=F, warning=F, fig.width = 7, fig.height=4.5, fig.align='center', message=F}
library(tidyverse)
library(Stat2Data)
data("PalmBeach")

PalmBeach %>%
  ggplot(aes(x = Bush, y = Buchanan)) +
  geom_point(size = 2.5) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
.question[
Why might the potential outlier be a problem?
]
]

---

### The impact of outliers

```{r echo=F, fig.width=7, fig.height=5, fig.align='center', message=F, warning=F}
lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))

PalmBeach %>%
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach)) %>%
  ggplot(aes(x = Bush)) +
  geom_point(aes(y = Buchanan), 
             size = 2.5) +
  geom_line(aes(y = pred_1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred_2), color = "red",
            lwd = 1.2, lty=2) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
* Blue, solid line: potential outlier is included in the data
* Red, dashed line: potential outlier is excluded
]

---

### Influential points

.large[
Outliers which substantially change the fitted line are called **influential.**
]

```{r echo=F, fig.width=7, fig.height=4.5, fig.align='center', message=F, warning=F}
lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))

PalmBeach %>%
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach)) %>%
  ggplot(aes(x = Bush)) +
  geom_point(aes(y = Buchanan), 
             size = 2.5) +
  geom_line(aes(y = pred_1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred_2), color = "red",
            lwd = 1.2, lty=2) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
.question[
Why would an outlier change the fitted line?
]
]

---

### Impact of influential points

.large[
$\widehat{\beta}_0$ and $\widehat{\beta}_1$ minimize $\sum \limits_{i=1}^n (y_i - \widehat{\beta}_0 - \widehat{\beta}_1 x_i)^2$
]

```{r echo=F, fig.width=7, fig.height=5, fig.align='center', message = F, warning=F}
lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))

PalmBeach %>%
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach)) %>%
  ggplot(aes(x = Bush)) +
  geom_segment(aes(x = 152846, y = 796.81,
                   xend = 152846, yend = 3407)) +
  geom_point(aes(y = Buchanan), 
             size = 2.5) +
  geom_line(aes(y = pred_1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred_2), color = "red",
            lwd = 1.2, lty=2) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

---

### Potential for influence

.large[
.question[
Which point do you think is more influential?
]
]

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(1)

x  = sample(1:100, 15)
y = x + rnorm(15, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)

x  = sample(1:100, 15)
y = x + rnorm(15, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 100, y = 30), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.large[
.abox[
(A) The one on the left
]

.bbox[
(B) The one on the right
]
]

---

### Potential for influence

.large[
Influence depends on how far $x$ is from $\overline{x}$.
]

.pull-left[
<ins> Some influence </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)

x  = sample(1:100, 15)
y = x + rnorm(15, sd=10)

x1 <- c(x, 50)
y1 <- c(y, 120)
lm1 <- lm(y ~ x)
lm2 <- lm(y1 ~ x1)

data.frame(x = x,
           y = y,
           pred1 = predict(lm1),
           pred2 = predict(lm2, newdata = data.frame(x1 = x, y1 = y))) %>%
  ggplot(aes(x = x)) +
  geom_point(aes(y = y), size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  geom_line(aes(y = pred1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred2), color = "red",
            lwd = 1.2, lty=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
<ins> More influence </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}

x1 <- c(x, 100)
y1 <- c(y, 30)
lm1 <- lm(y ~ x)
lm2 <- lm(y1 ~ x1)

data.frame(x = x,
           y = y,
           pred1 = predict(lm1),
           pred2 = predict(lm2, newdata = data.frame(x1 = x, y1 = y))) %>%
  ggplot(aes(x = x)) +
  geom_point(aes(y = y), size = 2.5) +
  geom_point(aes(x = 100, y = 30), color="red",
             size = 2.5) +
  geom_line(aes(y = pred1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred2), color = "red",
            lwd = 1.2, lty=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Influential points

.large[
.question[
Which point do you think is more influential?
]
]

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)

x  = sample(1:100, 15)
y = x + rnorm(15, sd=10)


data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)

x  = sample(1:100, 15)
y = x + rnorm(15, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 100, y = 100), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.large[
.abox[
(A) The one on the left
]

.bbox[
(B) The one on the right
]
]

---

### Influential points

.large[
Influence depends on both $x$ and $y$.
]

.pull-left[
<ins> Some influence </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)

x1 <- c(x, 50)
y1 <- c(y, 120)
lm1 <- lm(y ~ x)
lm2 <- lm(y1 ~ x1)

data.frame(x = x,
           y = y,
           pred1 = predict(lm1),
           pred2 = predict(lm2, newdata = data.frame(x1 = x, y1 = y))) %>%
  ggplot(aes(x = x)) +
  geom_point(aes(y = y), size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  geom_line(aes(y = pred1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred2), color = "red",
            lwd = 1.2, lty=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
<ins> Not influential </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
x1 <- c(x, 100)
y1 <- c(y, 100)
lm1 <- lm(y ~ x)
lm2 <- lm(y1 ~ x1)

data.frame(x = x,
           y = y,
           pred1 = predict(lm1),
           pred2 = predict(lm2, newdata = data.frame(x1 = x, y1 = y))) %>%
  ggplot(aes(x = x)) +
  geom_point(aes(y = y), size = 2.5) +
  geom_point(aes(x = 100, y = 100), color="red",
             size = 2.5) +
  geom_line(aes(y = pred1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred2), color = "red",
            lwd = 1.2, lty=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Assessing outliers: studentized residuals

.large[
**Studentized residuals:** $\dfrac{y_i - \widehat{y}_i}{\widehat{\sigma}_{\varepsilon}}$

where $\widehat{\sigma}_{\varepsilon}$ is computed without observation $i$. 

* Measure how many *standard deviations* away the observation is from the predicted value

.question[
What is a large number of standard deviations?
]
]

---

### Studentized residual plot

.large[
**Studentized residuals:** $\dfrac{y_i - \widehat{y}_i}{\widehat{\sigma}_{\varepsilon}}$

**Studentized residual plot:** plot studentized residuals on vertical axis, and $\widehat{y}$ on horizontal axis. Add lines at -3, -2, 2, and 3.
]

```{r echo=F, fig.width=7, fig.height=4.5, fig.align='center'}
vote_lm <- lm(Buchanan ~ Bush, data = PalmBeach)

PalmBeach %>%
  mutate(studentized_resids = rstudent(vote_lm),
         yhat = predict(vote_lm)) %>%
  ggplot(aes(x = yhat, y = studentized_resids)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd=1.2) +
  geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
              color = "orange", lty = 2, lwd=1.2) +
  labs(x = "Predicted votes for Buchanan",
       y = "Studentized residuals") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

---

### Studentized residual plot

.large[
**Studentized residual plot:** plot studentized residuals on vertical axis, and $\widehat{y}$ on horizontal axis. Add lines at -3, -2, 2, and 3.
]

```{r echo=F, fig.width=7, fig.height=4.5, fig.align='center'}
vote_lm <- lm(Buchanan ~ Bush, data = PalmBeach)

PalmBeach %>%
  mutate(studentized_resids = rstudent(vote_lm),
         yhat = predict(vote_lm)) %>%
  ggplot(aes(x = yhat, y = studentized_resids)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd=1.2) +
  geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
              color = "orange", lty = 2, lwd=1.2) +
  labs(x = "Predicted votes for Buchanan",
       y = "Studentized residuals") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
Points $< -3$ or $> 3$ are suspected outliers.
]

---

### Handling outliers

.large[
.question[
Suppose we identify a suspected outlier. What should we do about it?
]
]

---

### Handling outliers

.large[
.question[
Suppose we identify a suspected outlier. What should we do about it?
]
]

.large[
If the observation is clearly a measurement error:
* Remove it

Otherwise:
* Fit line with and without point, see how the estimated line changes
* If the estimated line changes substantially, report both fits
]

---

### Class activity

.large[
[https://sta112-s26.github.io/class_activities/ca_11.html](https://sta112-s26.github.io/class_activities/ca_11.html)
]

.large[
* Exploring outliers
* Submit HTML at end of class
]

