---
title: Inference and assumptions
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Last time: Activity

.large[
Data on 116 sparrows. Want to make a 98% confidence interval.

```{r, eval=F}
qt(..., df = ..., lower.tail=FALSE)
```

.question[
What do I fill in to calculate the critical value $t^*$?
]
]

---

### Activity

.large[
Data on 116 sparrows. Want to make a 98% confidence interval.

```{r}
qt(0.01, df = 114, lower.tail=FALSE)
```

]

---

### Activity

.large[
```
            Estimate Std. Error 
(Intercept)  8.75465    1.39601  
Weight       1.31341    0.09756 
```

$t^* = 2.36$

.question[
What is my 98% confidence interval?
]
]

---

### Activity

.large[
```
            Estimate Std. Error 
(Intercept)  8.75465    1.39601  
Weight       1.31341    0.09756 
```

$t^* = 2.36$

$$1.31 \pm 2.36(0.098) = (1.08, 1.54)$$

.question[
How do I interpret this confidence interval?
]
]

---

### Activity

.large[
```
            Estimate Std. Error 
(Intercept)  8.75465    1.39601  
Weight       1.31341    0.09756 
```

$t^* = 2.36$

$$1.31 \pm 2.36(0.098) = (1.08, 1.54)$$

.question[
Your friend claims that there is a 2% chance that the true slope $\beta_1$ is greater than 1.54 or less than 1.08. Is your friend correct?
]
]

---

### Activity

.large[
$$1.31 \pm 2.36(0.098) = (1.08, 1.54)$$

.question[
Your friend claims that there is a 2% chance that the true slope $\beta_1$ is greater than 1.54 or less than 1.08. Is your friend correct?
]
]

.large[
No. The true slope $\beta_1$ is a fixed (though unknown) parameter. There is no randomness involved in $\beta_1$, we just don't know what it is.

This means that for any specific interval we calculate -- e.g., (1.08, 1.54) -- $\beta_1$ is either *in the interval* or it is *not in the interval*. There is no probability involved! 

Probability only becomes involved when we imagine repeating the process *many times*.
]

---

### Linear regression assumptions

.large[
* **Shape assumption:** the shape of the regression model is at least approximately correct

To do *inference* (confidence intervals and hypothesis tests) we need some other assumptions:

* Constant variance
* Normality (with zero mean)
* Randomness
* Independence
]

---

### Constant variance

.large[
.center[
$y = \beta_0 + \beta_1 x + \varepsilon$
]
]

.large[
**Assumption:** Variance of the noise $\varepsilon$ is the same for all values of the predictor $x$
]

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
library(tidyverse)
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Constant variance") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Non-constant variance") +
  theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

---

### Assessing shape and constant variance: residual plots

.large[
**Residual plot:** plot residuals $y - \widehat{y}$ on vertical axis, and predicted values $\widehat{y}$ on horizontal axis. Add a horizontal line at 0.
]

```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=4}
library(latex2exp)
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### Residual plots

```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```

.large[
Shape assumption is reasonable if: 
* residuals appear to be scattered randomly above and below 0
* no clear patterns in the residuals
]

---

### Residual plots

```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=4}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```

.large[
Constant variance assumption is reasonable if:
* the band of residuals has relatively constant width
]

---

### Normality (with zero mean)

.large[
**Linear regression model:**

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$
]
]

.large[
**Assumption:** The noise term $\varepsilon$ follows a <ins>normal</ins> distribution, with mean 0


Example of a normal distribution:
]

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)
data.frame(x = seq(-4, 4, 0.1),
           y = dnorm(seq(-4, 4, 0.1))) %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(lwd=1.2) +
  theme_classic() +
  labs(x = "x", y = "Density")
```

---

### Normal distributions

.large[
Parameterized by mean $\mu$ and variance $\sigma^2$:

.center[
$N(\mu, \sigma^2)$
]
]

.pull-left[
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)
data.frame(x = seq(-10, 10, 0.1),
           y = dnorm(seq(-10, 10, 0.1))) %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(lwd=1.5) +
  theme_classic() +
  labs(x = "x", y = "Density",
       title = "N(0, 1)") +
  theme(text = element_text(size=25))
```
]

.pull-right[
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)
data.frame(x = seq(-8, 12, 0.1),
           y = dnorm(seq(-8, 12, 0.1), 2, 3)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(lwd=1.5) +
  theme_classic() +
  labs(x = "x", y = "Density",
       title = "N(2, 9)") +
  theme(text = element_text(size=25))
```
]

---

### Assessing normality

.large[
**Quantile-quantile (QQ) plots:** quantiles of residuals on y-axis, quantiles of theoretical normal distribution on x-axis.

If residuals are normal, plot should look approximately linear.
]

.pull-left[

.large[
<ins>Residuals look normal</ins>
]

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)
temp_x = 1:50
temp_y = 1:50 + rnorm(50, sd=10)
temp_lm <- lm(temp_y ~ temp_x)

data.frame(res = residuals(temp_lm)) %>%
  ggplot(aes(sample = res)) +
  geom_qq(size = 2) +
  geom_qq_line(size = 1.5) +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```
]

.pull-right[

.large[
<ins>Residuals do not look normal</ins>
]

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)
temp_x = 1:50
temp_y = 1:50 + rexp(50, rate = 1)
temp_lm <- lm(temp_y ~ temp_x)

data.frame(res = residuals(temp_lm)) %>%
  ggplot(aes(sample = res)) +
  geom_qq(size = 2) +
  geom_qq_line(size = 1.5) +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```
]

---

### QQ plot examples

.pull-left[

.large[
<ins>Residuals look normal</ins>
]

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(5)
temp_x = seq(1, 50, 0.5)
temp_y = seq(1, 50, 0.5) + rnorm(99, sd=10)
temp_lm <- lm(temp_y ~ temp_x)

data.frame(res = residuals(temp_lm)) %>%
  ggplot(aes(sample = res)) +
  geom_qq(size = 2) +
  geom_qq_line(size = 1.5) +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```
]

.pull-right[

.large[
<ins>Residuals do not look normal</ins>
]

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(2)
temp_x = seq(1, 50, 0.5)
temp_y = seq(1, 50, 0.5) + runif(99)*10
temp_lm <- lm(temp_y ~ temp_x)

data.frame(res = residuals(temp_lm)) %>%
  ggplot(aes(sample = res)) +
  geom_qq(size = 2) +
  geom_qq_line(size = 1.5) +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```
]

---

### Formal assumptions for inference

.large[
Assumptions about the relationship between variables:

* Shape
* Constant variance
* Normality (with zero mean)

Assumptions about how the data were collected:

* Independence
* Randomness
]

---

### Independence

.large[
**Assumption:** Observations are independent -- one point falling above or below the line has no influence on the location of another point

Independence can't be checked with plots. We need to think about how the data were collected.
]

---

### Independence?

.large[
.pull-left[
**Scenario:** You collect a sample of 116 sparrows from Kent island. For each sparrow, you record their wing length and weight. Your data has 116 rows (one for each sparrow), and two columns.
]

.pull-right[
Is it reasonable to assume the observations are independent?

.abox[
(A) Yes
]

.bbox[
(B) No
]
]
]

---

### Independence?

.large[
.pull-left[
**Scenario:** You go to Kent island, and find a sparrow. For 116 consecutive days, you record its weight and wing length. Your data has 116 rows, and two columns.
]

.pull-right[
Is it reasonable to assume the observations are independent?

.abox[
(A) Yes
]

.bbox[
(B) No
]
]
]

---

### Randomness

.large[
**Assumption:** The data are obtained using a random process, such as a random sample or randomized experiment

.pull-left[
<ins>Randomness assumption is reasonable</ins>

We collect data on a random sample of 116 sparrows from Kent island. For each sparrow, we record weight and wing length.
]

.pull-right[
<ins>Randomness assumption is not reasonable</ins>

We collect data on the weight and wing length of sparrows from Kent island. To collect the data, I deliberately choose sparrows that are very heavy, with very short wings.
]
]

---

### Randomness

.large[
**Assumption:** The data are obtained using a random process, such as a random sample or randomized experiment

.pull-left[
<ins>Randomness assumption is reasonable</ins>

We collect data on a random sample of 116 sparrows from Kent island. For each sparrow, we record weight and wing length.

**Conclusions from the estimated regression line can be generalized to a population of interest.**
]

.pull-right[
<ins>Randomness assumption is not reasonable</ins>

We collect data on the weight and wing length of sparrows from Kent island. To collect the data, I deliberately choose sparrows that are very heavy, with very short wings.

**Conclusions from the estimated regression line do not generalize.**
]
]

---

### Randomness

.large[
**Assumption:** The data are obtained using a random process, such as a random sample or randomized experiment

* Assumption about data collection process
* Necessary for defining the scope of inference
* Can't be checked with plots
]

---

### Putting it all together

.large[
For a quantitative response $y$ and a single quantitative predictor $x$, the <ins>simple linear regression model</ins> is

.center[
$y = \beta_0 + \beta_1 x + \varepsilon$
]
]

.large[
where $\varepsilon \sim N(0, \sigma_{\varepsilon}^2)$, and the noise $\varepsilon$ are independent from one another.
]

.large[
When we fit the model with a sample $(x_1, y_1), ..., (x_n, y_n)$, we also assume the data are generated by a random process that allows us to generalize to the population of interest.
]

---

### Summary: checking assumptions for inference

.large[
* Shape and constant variance: residual plot
* normality: QQ plot
* independence and randomness: think about data generation and collection
]

---

### Class activity

.large[
Data on doctors and hospitals from a random sample of 53 US counties:

* `County`: the name and state of the county
* `MDs`: the number of medical doctors in the county
* `Hospitals`: number of hospitals in the county
* `Beds`: the number of beds in the hospital

[https://sta112-s26.github.io/class_activities/ca_14.html](https://sta112-s26.github.io/class_activities/ca_14.html)
]

---

### Class activity

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
library(Stat2Data)
data("CountyHealth")
health_lm <- lm(MDs ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted number of doctors",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the shape and constant variance assumptions look reasonable?
]
]

---

### Class activity

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted number of doctors",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
* Shape may be slightly violated
* Constant variance is definitely violated
]

---

### Class activity

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
CountyHealth %>%
  mutate(residuals = residuals(health_lm)) %>%
  ggplot(aes(sample = residuals)) +
  geom_qq(size=2) +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```

.large[
.question[
Does the normality assumption look reasonable?
]
]

---

### Class activity

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
CountyHealth %>%
  mutate(residuals = residuals(health_lm)) %>%
  ggplot(aes(sample = residuals)) +
  geom_qq(size=2) +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```

.large[
The normality assumption is definitely violated.
]

---

### Transformation

.large[
Trying a square root transformation of the response:
]

```{r, echo=F, message=F, fig.width=7, fig.height=5}
health_lm <- lm(sqrt(MDs) ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted sqrt(number of doctors)",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
.question[
How do the shape and constant variance assumptions look now?
]
]

---

### Transformation

.large[
QQ plot with the transformed data:
]

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
CountyHealth %>%
  mutate(residuals = residuals(health_lm)) %>%
  ggplot(aes(sample = residuals)) +
  geom_qq(size=2) +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=20))
```

.large[
.question[
How does the normality assumption look now?
]
]


---

### Class activity

.large[
.question[
Do the independence and randomness assumptions seem reasonable for the `CountyHealth` data?
]
]

---

### Class activity

.large[
.question[
Do the independence and randomness assumptions seem reasonable for the `CountyHealth` data?
]
]

.large[
Yes, both seem reasonable.

* Independence: reasonable to assume the number of hospitals/doctors for one county doesn't affect the number of hospitals/doctors for another county
* Randomness: we are told the data come from a random sample
]



