---
title: Hypothesis testing with multiple regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Last time: Class activity

```{r, eval=F}
anova(m1)
```

```
Response: Height
           Df Sum Sq Mean Sq F value    Pr(>F)    
Age         1 5005.2  5005.2 575.820 < 2.2e-16 ***
Sex         1  295.8   295.8  34.035 2.235e-08 ***
Age:Sex     1  227.2   227.2  26.136 7.599e-07 ***
Residuals 194 1686.3     8.7                      
```

```{r, eval=F}
anova(m2)
```

```
Response: Height
           Df Sum Sq Mean Sq F value    Pr(>F)    
Sex         1  480.7  480.67   13.99 0.0002412 ***
Residuals 196 6733.9   34.36                      
```

.large[
$F = \dfrac{\frac{1}{\# \ \text{parameters tested}}(SSE_{reduced} - SSE_{full})}{\frac{1}{n-p} SSE_{full}}$

]

---

### Class activity

.large[
.question[
What is the p-value for this test?
]

```{r}
pf(290.35, df1 = 2, df2 = 194, lower.tail=F)
```

The p-value is basically 0, so we have strong evidence that there is a relationship between Age and Height, after accounting for Sex.
]

---

```{r include=F}
library(Stat2Data)
library(tidyverse)
data(Kids198)
Kids198$Sex = 1 - Kids198$Sex
```


### Testing in R

.large[
We can do comparisons between nested models a little more efficiently:

```{r}
m1 <- lm(Height ~ Age*Sex, data = Kids198)
m2 <- lm(Height ~ Sex, data = Kids198)

anova(m2, m1)
```
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$
$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

.question[
How did we test a single coefficient in simple linear regression?
]
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$
$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

.question[
How did we test a single coefficient in simple linear regression?
]
]

.large[
With a t-test!
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

$t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}}$

Under $H_0$, this test statistic follows a $t$ distribution.

.question[
What do you think should be the degrees of freedom of the $t$ distribution?
]
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

$t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}}$

Under $H_0$, this test statistic follows a $t$ distribution.

.question[
What do you think should be the degrees of freedom of the $t$ distribution?
]
]

.large[
$n - p$
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

$t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}}$

Under $H_0$, this test statistic follows a $t_{n-p}$ distribution.

]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

```{r}
m1 <- lm(Height ~ Age*Sex, data = Kids198)
```

```
             Estimate Std. Error t value Pr(>|t|)    
(Intercept) 32.960916   1.537781  21.434  < 2e-16 ***
Age          0.182543   0.009383  19.454  < 2e-16 ***
Sex         -7.727846   2.034957  -3.798 0.000195 ***
Age:Sex      0.064134   0.012545   5.112  7.6e-07 ***
```

$t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}} = \dfrac{0.0641}{0.0125} = 5.112$

p-value = $7.6 \times 10^{-7}$
]


---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Research question:** Is the slope on Age *larger* for boys than for girls?

.question[
What hypotheses could I test to answer this research question?
]

]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Research question:** Is the slope on Age *larger* for boys than for girls?

.question[
What hypotheses could I test to answer this research question?
]

]

.large[
$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 > 0$

**Test statistic:** $t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}} = 5.112$

.question[
Which tail do I use to calculate the p-value?
]
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Research question:** Is the slope on Age *larger* for boys than for girls?

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 > 0$

**Test statistic:** $t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}} = 5.112$

```{r}
pt(5.112, df = 194, lower.tail=F)
```

p-value = $3.8 \times 10^{-7}$, so we have strong evidence that the slope on Age is larger for boys than for girls.
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Research question:** Is the slope on Age *larger* for boys than for girls?

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 > 0$

.question[
Can I use a nested F test here?
]

]

.large[
No -- a nested F test doesn't let me test a specific direction ( $\beta_3 > 0$ )
]

---

### t tests vs. nested F tests

.large[

.pull-left[
**t test:** test a single parameter

Some example hypotheses:
* $H_0: \beta_3 = 0 \hspace{0.5cm} H_A: \beta_3 \neq 0$
* $H_0: \beta_3 = 0 \hspace{0.5cm} H_A: \beta_3 > 0$
* $H_0: \beta_3 = 1 \hspace{0.5cm} H_A: \beta_3 < 1$
]

.pull-right[
**nested F test:** test one or more parameters

Some example hypotheses:

$H_0: \beta_1 = \beta_2 = 0$
$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$
]

]

---

### Confidence intervals

.large[
Just like with simple linear regression, we can construct confidence intervals for single parameters.

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

Confidence interval for $\beta_3$:
]

---

### Confidence intervals

.large[
Just like with simple linear regression, we can construct confidence intervals for single parameters.

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

Confidence interval for $\beta_3$:

$$\widehat{\beta}_3 \pm t^* SE_{\widehat{\beta}_3}$$

where $t^*$ is the critical value for a $t_{n-p}$ distribution.
]

---

### Confidence intervals

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Confidence interval:** $\hspace{1cm} \widehat{\beta}_3 \pm t^* SE_{\widehat{\beta}_3}$

.question[
Suppose I want to calculate a 99% confidence interval. How do I find the critical value $t^*$?
]
]

---

### Confidence intervals

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Confidence interval:** $\hspace{1cm} \widehat{\beta}_3 \pm t^* SE_{\widehat{\beta}_3}$

.question[
Suppose I want to calculate a 99% confidence interval. How do I find the critical value $t^*$?
]
]

.large[
```{r}
qt(0.005, df=194, lower.tail=F)
```

$t^* = 2.60$
]

---

### Confidence intervals

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Confidence interval:** $\hspace{1cm} \widehat{\beta}_3 \pm t^* SE_{\widehat{\beta}_3}$

$$0.0641 \pm 2.60 \cdot 0.0125 = (0.0316, 0.0966)$$

So we are 99% confident that a one-month increase in age is associated with an change in height that is between 0.0316 and 0.0966 inches greater for boys than for girls.
]






