---
title: Assumptions and diagnostics
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Predictions revisited

.large[
Here are two different datasets with similar correlation ($r \approx 0.8$):
]

.pull-left[
```{r, echo=F, message=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)
set.seed(3)
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = log(10*(1:100)) + rnorm(100, sd=0.3)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.large[
.question[
For each dataset, suppose I wish to make a prediction for $x = 2$. For which dataset would I expect to predict better?
]
]

---

### Predictions revisited

.large[
Here are two different datasets with similar correlation ($r \approx 0.8$):
]

.pull-left[
```{r, echo=F, message=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)
set.seed(3)
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = log(10*(1:100)) + rnorm(100, sd=0.3)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.large[
.question[
For each dataset, suppose I wish to make a prediction for $x = 70$. For which dataset would I expect to predict better?
]
]

---

### Predictions revisited

.large[
Here are two different datasets with similar correlation ($r \approx 0.8$):
]

.pull-left[
```{r, echo=F, message=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)
set.seed(3)
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = log(10*(1:100)) + rnorm(100, sd=0.3)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.large[
.question[
For which dataset does the line appear to be a better fit?
]
]

---

### Predictions and linearity

.large[
If we use a line to describe non-linear data, the quality of our predictions may depend on the value of $x$:
]

.pull-left[
```{r, echo=F, message=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)
set.seed(3)
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = log(10*(1:100)) + rnorm(100, sd=0.3)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 1.5, col="red") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))
```
]

---

### Predictions revisited

.large[
Here are two different datasets with similar correlation ($r \approx 0.8$):
]

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(3)

data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 10.5, col="red") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 10.5, col="red") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

.large[
.question[
For each dataset, suppose I wish to make a prediction for $x = 10$. For which dataset would I expect to predict better?
]
]

---

### Predictions revisited

.large[
Here are two different datasets with similar correlation ($r \approx 0.8$):
]

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(3)

data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = 70.5, col="red") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

.large[
.question[
For each dataset, suppose I wish to make a prediction for $x = 70$. For which dataset would I expect to predict better?
]
]

---

### Predictions revisited

.large[
Here are two different datasets with similar correlation ($r \approx 0.8$):
]

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(3)

data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

.large[
.question[
Does a linear fit seem appropriate for both datasets?
]
]

---

### Predictions and variability

.large[
If the variability of our residuals changes for different values of $x$, the quality of our predictions depends on $x$:
]

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
set.seed(3)

data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = c(10.5, 70.5), col="red") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=4, message=F, warning=F}
data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)*(1:100)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  geom_vline(xintercept = c(10.5, 70.5), col="red") +
  theme(text = element_text(size=25)) +
  scale_y_continuous(limits=c(-50, 150))
```
]

---

### Common linear model assumptions

.large[
$$Y = \beta_0 + \beta_1 X + \varepsilon$$

**Shape assumption:** the shape of the regression model is at least approximately correct
* When we fit linear regression, we assume the relationship is approximately linear

**Constant variance assumption:** Variance of the noise $\varepsilon$ is the same for all values of the predictor $X$
]

---

### Assessing shape and constant variance: residual plots

.large[
**Residual plot:** plot residuals $y - \widehat{y}$ on vertical axis, and predicted values $\widehat{y}$ on horizontal axis. Add a horizontal line at 0.
]

```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=4}
library(latex2exp)
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### Residual plots

```{r, echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
library(gridExtra)
set.seed(3)

d1 <- data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20))

p1 <- d1 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))

d2 <- data.frame(x = 1:100,
           y = log(10*(1:100)) + rnorm(100, sd=0.3)) 

p2 <- d2 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))

m1 <- lm(y ~ x, data = d1)
m2 <- lm(y ~ x, data = d2)

p3 <- d1 |>
  mutate(pred = predict(m1), 
         residuals = resid(m1)) |>
  ggplot(aes(x = pred, y = residuals)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color = "blue") +
  theme_bw() +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        text = element_text(size = 25)) +
  labs(x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$"))

p4 <- d2 |>
  mutate(pred = predict(m2), 
         residuals = resid(m2)) |>
  ggplot(aes(x = pred, y = residuals)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color = "blue") +
  theme_bw() +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        text = element_text(size = 25)) +
  labs(x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$"))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

.large[
Shape assumption is reasonable if: 
* residuals appear to be scattered randomly above and below 0
* no clear patterns in the residuals
]

---

### Residual plots

```{r, echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
library(gridExtra)
set.seed(3)

d1 <- data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20))

p1 <- d1 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Dataset 1") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))

d2 <- data.frame(x = 1:100,
           y = 1:100 + rnorm(100, sd=20)*(1:100)*0.02)

p2 <- d2 %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Dataset 2") +
  theme_bw() +
  geom_smooth(se=F, method="lm") +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        text = element_text(size=25))

m1 <- lm(y ~ x, data = d1)
m2 <- lm(y ~ x, data = d2)

p3 <- d1 |>
  mutate(pred = predict(m1), 
         residuals = resid(m1)) |>
  ggplot(aes(x = pred, y = residuals)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color = "blue") +
  theme_bw() +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        text = element_text(size = 25)) +
  labs(x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$"))

p4 <- d2 |>
  mutate(pred = predict(m2), 
         residuals = resid(m2)) |>
  ggplot(aes(x = pred, y = residuals)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color = "blue") +
  theme_bw() +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        text = element_text(size = 25)) +
  labs(x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$"))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

.large[
Constant variance assumption is reasonable if:
* the band of residuals has relatively constant width
]

---

### Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)*(100:1)*0.02) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.pull-right[
.large[
Which assumption looks reasonable?

.abox[
(A) Shape
]

.bbox[
(B) Constant variance
]

.cbox[
(C) Both shape and constant variance
]

.dbox[
(D) Neither shape nor constant variance
]
]
]


---

### Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
sim_dat <- data.frame(x = seq(-50, 50),
           y = 0.02*seq(-50, 50)^2 + rnorm(101, sd=10)*(1:101)*0.02)
sim_reg <- lm(y ~ x, data = sim_dat)

data.frame(pred = sim_reg$fitted.values,
           res = sim_reg$residuals) %>%
  ggplot(aes(x = pred, y = res)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.pull-right[
.large[
Which assumption looks reasonable?

.abox[
(A) Shape
]

.bbox[
(B) Constant variance
]

.cbox[
(C) Both shape and constant variance
]

.dbox[
(D) Neither shape nor constant variance
]
]
]


---

### Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
set.seed(2)

sim_dat <- data.frame(x = seq(-50, 50),
           y = 0.02*seq(-50, 50)^2 + rnorm(101, sd=10))
sim_reg <- lm(y ~ x, data = sim_dat)

data.frame(pred = sim_reg$fitted.values,
           res = sim_reg$residuals) %>%
  ggplot(aes(x = pred, y = res)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.pull-right[
.large[
Which assumption looks reasonable?

.abox[
(A) Shape
]

.bbox[
(B) Constant variance
]

.cbox[
(C) Both shape and constant variance
]

.dbox[
(D) Neither shape nor constant variance
]
]
]


---

### Concept check

.pull-left[
```{r, echo = F, message=F, fig.align='center', fig.width=6, fig.height=6}
data.frame(x = 1:100,
           y = rnorm(100, sd=20)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              color = "blue", lwd=1.2) +
  labs(title = "Residual plot",
       x = TeX("$\\widehat{y}$"),
       y = TeX("$y - \\widehat{y}$")) +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.pull-right[
.large[
Which assumption looks reasonable?

.abox[
(A) Shape
]

.bbox[
(B) Constant variance
]

.cbox[
(C) Both shape and constant variance
]

.dbox[
(D) Neither shape nor constant variance
]
]
]

---

### Class activity

.large[
[https://sta112-s26.github.io/class_activities/ca_08.html](https://sta112-s26.github.io/class_activities/ca_08.html)
]

.large[
* Evaluate shape and constant variance assumptions with residual plots
* Submit HTML at end of class
]

---

### Class activity

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
library(Stat2Data)
data("CountyHealth")
health_lm <- lm(MDs ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted number of doctors",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the shape and constant variance assumptions look reasonable?
]
]




