---
title: Nonlinear relationships and transformations
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Last time

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
library(Stat2Data)
library(tidyverse)
data("CountyHealth")
health_lm <- lm(MDs ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted number of doctors",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the shape and constant variance assumptions look reasonable?
]
]

---

### Today

```{r, echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
library(Stat2Data)
library(tidyverse)
data("CountyHealth")
health_lm <- lm(MDs ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted number of doctors",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
Today: handling non-linear relationships
]

---

### Example

```{r, echo=F, message=F, warning=F, fig.width=10, fig.height=5}
library(gridExtra)

set.seed(4)
n <- 100
x <- seq(1, 20, length.out = n)
eps <- rnorm(n, sd=0.25)

y <- log(x) + eps
p1 <- data.frame(x = x, y = y) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(title = "Original data") +
  theme(text = element_text(size = 15))

m1 <- lm(y ~ x)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))

grid.arrange(p1, p2, ncol=2)
```

.large[
.question[
The relationship between $x$ and $y$ does not appear linear. How could we model that relationship instead?
]
]

---

### Example: log transform

.large[
$$Y = \beta_0 + \beta_1 \log(X) + \varepsilon$$
]

```{r, echo=F, message=F, warning=F, fig.width=7, fig.height=6, fig.align='center'}
library(gridExtra)

set.seed(4)
n <- 100
x <- seq(1, 20, length.out = n)
eps <- rnorm(n, sd=0.25)

y <- log(x) + eps

m1 <- lm(y ~ x)
m2 <- lm(y ~ log(x))

p1 <- data.frame(x = x, y = y) |>
  ggplot() +
  geom_point(aes(x = x, y = y)) +
  geom_smooth(aes(x = x, y = y), se=F, method="lm") +
  geom_line(aes(x = x, y = predict(m2)), color="red") +
  theme_bw() +
  labs(title = "Original data") +
  theme(text = element_text(size = 15))

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))

p3 <- data.frame(x = x, y = y) |>
  ggplot(aes(x = log(x), y = y)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(title = "Log transform") +
  theme(text = element_text(size = 15))

p4 <- data.frame(preds = predict(m2), 
                 resids = residuals(m2)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))


grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Example

```{r, echo=F, message=F, warning=F, fig.width=10, fig.height=5}
library(gridExtra)

set.seed(4)
n <- 100
x <- seq(1, 20, length.out = n)
eps <- rnorm(n, sd=0.25)

y <- 0.01*x^2 + eps
p1 <- data.frame(x = x, y = y) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(title = "Original data") +
  theme(text = element_text(size = 15))

m1 <- lm(y ~ x)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))

grid.arrange(p1, p2, ncol=2)
```

.large[
.question[
The relationship between $x$ and $y$ does not appear linear. How could we model that relationship instead?
]
]

---

### Example: squaring X

.large[
$$Y = \beta_0 + \beta_1 X^2 + \varepsilon$$
]

```{r, echo=F, message=F, warning=F, fig.width=7, fig.height=6, fig.align='center'}
library(gridExtra)

set.seed(4)
n <- 100
x <- seq(1, 20, length.out = n)
eps <- rnorm(n, sd=0.25)

y <- 0.01*x^2 + eps

m1 <- lm(y ~ x)
m2 <- lm(y ~ I(x^2))

p1 <- data.frame(x = x, y = y) |>
  ggplot() +
  geom_point(aes(x = x, y = y)) +
  geom_smooth(aes(x = x, y = y), se=F, method="lm") +
  geom_line(aes(x = x, y = predict(m2)), color="red") +
  theme_bw() +
  labs(title = "Original data") +
  theme(text = element_text(size = 15))

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))

p3 <- data.frame(x = x, y = y) |>
  ggplot(aes(x = x^2, y = y)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(title = "Square transform") +
  theme(text = element_text(size = 15))

p4 <- data.frame(preds = predict(m2), 
                 resids = residuals(m2)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))


grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Example

```{r, echo=F, message=F, warning=F, fig.width=10, fig.height=5}
library(gridExtra)

set.seed(4)
n <- 100
x <- seq(1, 20, length.out = n)
eps <- rnorm(n, sd=0.25)

y <- 5/x + eps
p1 <- data.frame(x = x, y = y) |>
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(title = "Original data") +
  theme(text = element_text(size = 15))

m1 <- lm(y ~ x)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))

grid.arrange(p1, p2, ncol=2)
```

.large[
.question[
The relationship between $x$ and $y$ does not appear linear. How could we model that relationship instead?
]
]

---

### Example: reciprocal transform

.large[
$$Y = \beta_0 + \beta_1 \ (1/X) + \varepsilon$$
]

```{r, echo=F, message=F, warning=F, fig.width=7, fig.height=6, fig.align='center'}
library(gridExtra)

set.seed(4)
n <- 100
x <- seq(1, 20, length.out = n)
eps <- rnorm(n, sd=0.25)

y <- 5/x + eps

m1 <- lm(y ~ x)
m2 <- lm(y ~ I(1/x))

p1 <- data.frame(x = x, y = y) |>
  ggplot() +
  geom_point(aes(x = x, y = y)) +
  geom_smooth(aes(x = x, y = y), se=F, method="lm") +
  geom_line(aes(x = x, y = predict(m2)), color="red") +
  theme_bw() +
  labs(title = "Original data") +
  theme(text = element_text(size = 15))

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))

p3 <- data.frame(x = x, y = y) |>
  ggplot(aes(x = 1/x, y = y)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(title = "Reciprocal transform") +
  theme(text = element_text(size = 15))

p4 <- data.frame(preds = predict(m2), 
                 resids = residuals(m2)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predictions", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 15))


grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Choosing transformations

.large[
To model nonlinear relationships, try transformations for $x$ that looks like the relationship
]

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=5.5}
library(patchwork)
p1 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = sqrt(x))) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Square root")

p2 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = x^2)) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Square")

p3 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = log(x))) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Log")

p4 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = 1/x)) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Reciprocal")

(p1 + p2)/(p3 + p4)
```

---

### Example

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=4}
library(gapminder)

gapminder_2007 <- gapminder |>
  filter(year == 2007)

p1 <- gapminder_2007 |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  theme_bw() +
  labs(x = "GDP per capita ($)", y = "Life Expectancy (years)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

m1 <- lm(lifeExp ~ gdpPercap, data = gapminder_2007)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted life expectancy", y = "Residuals") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```

.large[
.question[
What transformation could I try for this data?
]
]

---

### Log transformation

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=7}
p1 <- gapminder_2007 |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  theme_bw() +
  labs(x = "GDP per capita ($)", y = "Life Expectancy (years)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

m1 <- lm(lifeExp ~ gdpPercap, data = gapminder_2007)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted life expectancy", y = "Residuals") +
  theme(text = element_text(size = 20))

m2 <- lm(lifeExp ~ log(gdpPercap), data = gapminder_2007)

p3 <- gapminder_2007 |>
  ggplot(aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point() +
  theme_bw() +
  labs(x = "log(GDP per capita)", y = "Life Expectancy (years)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

p4 <- data.frame(preds = predict(m2), 
                 resids = residuals(m2)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted life expectancy", y = "Residuals") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Log transformation

```{r echo=F, message=F, fig.align='center', fig.width=6, fig.height=4}
p3
```


.large[
$$\widehat{\text{LifeExpectancy}} = 4.95 + 7.20 \log(\text{GDP per capita})$$

.question[
What life expectancy would I predict for a GDP per capita of $20000?
]
]

---

### Example

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=4}
set.seed(4)

n <- 100
x <- seq(1, 3, length.out = n)
eps <- rnorm(n, sd=0.25)

y <- 30000*exp(x + eps)
x <- x*10

ex_df <- data.frame(SqFt = x, Price = y)

p1 <- ex_df |>
  ggplot(aes(x = SqFt, y = Price)) +
  geom_point() +
  theme_bw() +
  labs(x = "Area (hundreds of sq ft)", y = "Price ($)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

m1 <- lm(Price ~ SqFt, data = ex_df)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted price", y = "Residuals") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```

.large[
.question[
What transformation could I try for this data?
]
]

---

### Trying squaring X

.large[
Model: $Price = \beta_0 + \beta_1 (Area)^2 + \varepsilon$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=6}
p1 <- ex_df |>
  ggplot(aes(x = SqFt, y = Price)) +
  geom_point() +
  theme_bw() +
  labs(x = "Area (hundreds of sq ft)", y = "Price ($)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

m1 <- lm(Price ~ SqFt, data = ex_df)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted price", y = "Residuals") +
  theme(text = element_text(size = 20))

m2 <- lm(Price ~ I(SqFt^2), data = ex_df)

p3 <- ex_df |>
  ggplot(aes(x = SqFt^2, y = Price)) +
  geom_point() +
  theme_bw() +
  labs(x = "(Area)^2", y = "Price ($)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

p4 <- data.frame(preds = predict(m2), 
                 resids = residuals(m2)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted price", y = "Residuals") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Trying squaring X

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=4}
grid.arrange(p3, p4, ncol=2)
```

.large[
.question[
What do you notice about the residual plot after trying the transformation?
]
]

---

### Another attempt: transforming the response!

.large[
Model: $\log(Price) = \beta_0 + \beta_1 Area + \varepsilon$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=6}
p1 <- ex_df |>
  ggplot(aes(x = SqFt, y = Price)) +
  geom_point() +
  theme_bw() +
  labs(x = "Area (hundreds of sq ft)", y = "Price ($)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

m1 <- lm(Price ~ SqFt, data = ex_df)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted price", y = "Residuals") +
  theme(text = element_text(size = 20))

m2 <- lm(log(Price) ~ SqFt, data = ex_df)

p3 <- ex_df |>
  ggplot(aes(x = SqFt, y = log(Price))) +
  geom_point() +
  theme_bw() +
  labs(x = "Area (hundreds of sq ft)", y = "log(Price)") +
  geom_smooth(se=F, method="lm") +
  theme(text = element_text(size = 20))

p4 <- data.frame(preds = predict(m2), 
                 resids = residuals(m2)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted log(price)", y = "Residuals") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Log transformed response

.large[
$$\widehat{\log(\text{Price})} = 10.51 + 0.091 \ \text{Area}$$

* Here Price is in $, Area is in hundreds of square feet
]

.large[
.question[
How would I calculate the predicted *price* (in $) for a house of 2000 square feet?
]
]

---

### Class activity


.large[
Work on the handout, then we will discuss as a class
]

---

### Class activity

.large[
$$\widehat{\text{LifeExpectancy}} = 4.95 + 7.20 \log(\text{GDP per capita})$$

.question[
What is the predicted price for GDP per capita of $20200?
]
]

---

### Class activity

.large[
$$\widehat{\text{LifeExpectancy}} = 4.95 + 7.20 \log(\text{GDP per capita})$$

* When GDP per capita = 20000: $\widehat{\text{LifeExpectancy}} = 76.255$
* When GDP per capita = 20200: $\widehat{\text{LifeExpectancy}} = 76.327$

Change:
]

---

### Class activity

.large[
$$\widehat{\log(\text{Price})} = 10.51 + 0.091 \ \text{Area}$$

* Price in $, Area in hundres of sq ft
]

.large[
.question[
What is the predicted price for a house of 2100 square feet?
]
]

---

### Class activity

.large[
$$\widehat{\log(\text{Price})} = 10.51 + 0.091 \ \text{Area}$$

* Price in $, Area in hundres of sq ft
* When Area = 20: $\widehat{\text{Price}} = 226386$
* When Area = 21: $\widehat{\text{Price}} = 247954$

Change:
]

---

### Class activity

.large[
$$\widehat{\log(\text{Price})} = 10.51 + 0.091 \ \text{Area}$$

* Price in $, Area in hundres of sq ft
* When Area = 20: $\widehat{\text{Price}} = 226386$
* When Area = 21: $\widehat{\text{Price}} = 247954$

Change: $\dfrac{247954}{226386} = 1.095 = e^{0.091}$
]













