---
title: Nonlinear relationships and transformations
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### More transformations


```{r, echo=F, message=F, warning=F, fig.width=10, fig.height=5}
library(tidyverse)
library(Stat2Data)
library(gridExtra)

data("SpeciesArea")

p1 <- SpeciesArea |>
  ggplot(aes(x = Area, y = Species)) +
  geom_point() +
  theme_bw() +
  labs(x = "Area (sq km)", y = "Number of species") +
  theme(text = element_text(size = 20)) +
  geom_smooth(se=F, method="lm")

m1 <- lm(Species ~ Area, data = SpeciesArea)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted number of species", y = "Residuals", 
       title = "Residual plot") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```

.large[
.question[
What transformations could we try here?
]
]

---

### Trying some transformations

```{r, echo=F, message=F, fig.width=8, fig.height=7, fig.align='center'}
p1 <- SpeciesArea |>
  ggplot(aes(x = log(Area), y = Species)) +
  geom_point() +
  theme_bw() +
  labs(x = "log(Area)", y = "Number of species") +
  theme(text = element_text(size = 20)) +
  geom_smooth(se=F, method="lm")

m1 <- lm(Species ~ log(Area), data = SpeciesArea)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted number of species", y = "Residuals") +
  theme(text = element_text(size = 20))


p3 <- SpeciesArea |>
  ggplot(aes(x = sqrt(Area), y = Species)) +
  geom_point() +
  theme_bw() +
  labs(x = "sqrt(Area)", y = "Number of species") +
  theme(text = element_text(size = 20)) +
  geom_smooth(se=F, method="lm")

m1 <- lm(Species ~ sqrt(Area), data = SpeciesArea)

p4 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted number of species", y = "Residuals") +
  theme(text = element_text(size = 20))



grid.arrange(p1, p2, p3, p4, ncol=2)
```

---

### Transform both predictor and response?

.large[
$$\log(Species) = \beta_0 + \beta_1 \log(Area) + \varepsilon$$
]

```{r, echo=F, message=F, warning=F, fig.width=10, fig.height=5}
p1 <- SpeciesArea |>
  ggplot(aes(x = log(Area), y = log(Species))) +
  geom_point() +
  theme_bw() +
  labs(x = "log(Area)", y = "log(Number of species)") +
  theme(text = element_text(size = 20)) +
  geom_smooth(se=F, method="lm")

m1 <- lm(log(Species) ~ log(Area), data = SpeciesArea)

p2 <- data.frame(preds = predict(m1), 
                 resids = residuals(m1)) |>
  ggplot(aes(x = preds, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, intercept = 0, color="blue") +
  theme_bw() +
  labs(x = "Predicted number of species", y = "Residuals") +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```

---

### Interpreting the fitted model

.large[
$$\widehat{\log(Species)} = 1.625 + 0.246 \log (Area)$$
]

---

### Summary of model interpretation so far

$\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 x$

* A one-unit increase in $x$ is associated with a change in $y$ of $\widehat{\beta}_1$, on average

$\widehat{y} = \widehat{\beta}_0 + \widehat{\beta}_1 log(x)$

* A 1% increase in $x$ is associated with a change in $y$ of $\widehat{\beta}_1/100$, on average

$\widehat{log(y)} = \widehat{\beta}_0 + \widehat{\beta}_1 x$

* A one-unit increase in $x$ is associated with a change in $y$ by a **factor** of $e^{\widehat{\beta}_1}$, on average

$\widehat{log(y)} = \widehat{\beta}_0 + \widehat{\beta}_1 log(x)$

* A 1% increase in $x$ is associated with a change in $y$ by a **factor** of $(1.01)^{\widehat{\beta}_1}$, on average


---

### Kentucky Derby data

.large[
Data on the Kentucky Derby winner from each year between 1875 and 2012. Variables include

* `year`: year of race
* `speed`: speed of winning horse (mph)
* `time`: winning time (seconds)
* `condition`: condition of the track on the day of the race
]

---

### EDA

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
derby <- read.delim("KYDerby17.txt")

derby %>%
  ggplot(aes(x = Year, y = Time)) +
  geom_point(size = 2.5) +
  geom_line() +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
What do you notice about the winning times?
]
]

---

### EDA

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
derby <- read.delim("KYDerby17.txt")

derby %>%
  ggplot(aes(x = Year, y = Time)) +
  geom_point(size = 2.5) +
  geom_line() +
  theme_bw() +
  theme(text = element_text(size = 25))

derby <- derby |> filter(Year > 1900)
```

.large[
.question[
What do you notice about the winning times?
]
]

.large[
There's a big drop around 1900, then they decrease gradually.
]

---

### EDA

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

derby %>%
  ggplot(aes(x = Year, y = speed)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
How would you describe the relationship?
]
]

---

### Fitting a linear regression

.large[
$$\text{Speed} = \beta_0 + \beta_1 \ \text{Year} + \varepsilon$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=5}

p1 <- derby %>%
  ggplot(aes(x = Year, y = speed)) +
  geom_point(size = 2.5) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  theme(text = element_text(size = 20))

derby_lm <- lm(speed ~ Year, data = derby)

p2 <- derby %>%
  mutate(pred = predict(derby_lm),
         resid = residuals(derby_lm)) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, color = "blue", lwd=1.2) +
  labs(x = "Predicted speed", y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 20))

grid.arrange(p1, p2, ncol=2)
```

---

### Polynomial transformations

.large[
Another possible transformation is a **polynomial**.

.question[
**Definition:** A polynomial transformation of order $k$ is a linear combination of the terms $X$, $X^2$, ..., $X^k$:
$$Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \cdots + \beta_k X^k + \varepsilon$$
]
]

.large[
* Example: If we use a polynomial of order 3, our model is 
$$Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \varepsilon$$
]

---

### Polynomial transformations

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

derby %>%
  ggplot(aes(x = Year, y = speed)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
What order polynomial do you think we should use?
]
]

---

### Polynomial transformation

.large[
Let's try a polynomial of order 2:

$$\text{Speed} = \beta_0 + \beta_1 \ \text{Year} +  \beta_2 \ \text{Year}^2 + \varepsilon$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

derby %>%
  ggplot(aes(x = Year, y = speed)) +
  geom_point(size = 2.5) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

---

### Diagnostic plots

.large[
We use residual plots to check the model assumptions after transforming our predictor, just like with other transformations.
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

derby_lm <- lm(speed ~ I(Year) + I(Year^2), data = derby)

derby %>%
  mutate(pred = predict(derby_lm),
         resid = residuals(derby_lm)) %>%
  ggplot(aes(x = pred, y = resid)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, color = "blue", lwd=1.2) +
  labs(x = "Predicted speed", y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
.question[
How does the shape assumption look now?
]
]

---

### Fitting the model in R

.large[
```{r eval=F}
derby_lm <- lm(speed ~ poly(Year, 2, raw=TRUE), 
               data = derby)
```
]

.large[
* `poly` means "fit a polynomial"
* `poly(..., 2, ...)` specifies a second-order polynomial (quadratic)
]

---

### Fitting the model in R

.large[
```{r}
derby_lm <- lm(speed ~ poly(Year, 2, raw=TRUE), 
               data = derby)
derby_lm
```
]

.large[
$$\widehat{\text{Speed}} = -1120 + 1.16 \ \text{Year} - 0.00029 \ \text{Year}^2$$
]

---

### Interpreting the fitted model

.large[
$$\widehat{\text{Speed}} = -1120 + 1.16 \ \text{Year} - 0.00029 \ \text{Year}^2$$
]

.large[
.question[
How would I interpret the intercept of the fitted polynomial regression model?
]
]

---

### Interpreting the fitted model

.large[
$$\widehat{\text{Speed}} = -1120 + 1.16 \ \text{Year} - 0.00029 \ \text{Year}^2$$
]

.large[
.question[
How would I interpret the intercept of the fitted polynomial regression model?
]
]

.large[
If Year = 0, we predict a speed of -1120 mph (doesn't make much sense, since year is nowhere near 0)
]

---

### Interpreting the fitted model

.large[
$$\widehat{\text{Speed}} = -1120 + 1.16 \ \text{Year} - 0.00029 \ \text{Year}^2$$
]

.large[
.question[
How would I interpret the terms on $\text{Year}$ and $\text{Year}^2$?
]
]

---

### Interpreting the fitted model

.large[
$$\widehat{\text{Speed}} = -1120 + 1.16 \ \text{Year} - 0.00029 \ \text{Year}^2$$
]

.large[
.question[
How would I interpret the terms on $\text{Year}$ and $\text{Year}^2$?
]
]

.large[
They're kind of difficult to interpret -- if I change $\text{Year}$, I also change $\text{Year}^2$
]

---

### Interpreting the fitted model

.large[
$$\widehat{\text{Speed}} = -1120 + 1.16 \ \text{Year} - 0.00029 \ \text{Year}^2$$
]

.large[
.question[
Can I interpret the *sign* on $\text{Year}^2$?
]
]

---

### Interpreting the fitted model

.large[
$$\widehat{\text{Speed}} = -1120 + 1.16 \ \text{Year} - 0.00029 \ \text{Year}^2$$
]

.large[
.question[
Can I interpret the *sign* on $\text{Year}^2$?
]
]

.large[
Yes (for a polynomial of order 2 specifically). The sign on $\text{Year}^2$ tells us about the concavity of the relationship.
]

---

### Class activity

.large[
[https://sta112-s26.github.io/class_activities/ca_10.html](https://sta112-s26.github.io/class_activities/ca_10.html)
]

.large[
* Experiment with polynomial regression
* Submit HTML at end of class
]

---

### Previously: polynomial transformations

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
library(Stat2Data)
data("CO2Germany")
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$$\widehat{\text{CO2}} = 414.975 - 0.476 \ \text{Day} + 0.0012 \ \text{Day}^2$$
]

---

### Previously: polynomial transformations

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
library(Stat2Data)
data("CO2Germany")
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$$\widehat{\text{CO2}} = 414.975 - 0.476 \ \text{Day} + 0.0012 \ \text{Day}^2$$

.question[
Can I use the correlation to calculate the relationship between Day and CO2?
]
]

---

### Measuring variability in the response

.large[
* When we fit a regression model with least squares, we choose parameters to minimize the sum of squared residuals:

$$SSE = \sum_i (y_i - \widehat{y}_i)^2$$
]

.large[
* We can think of SSE as the "amount of variability" in $Y$ after predicting with our regression model
* The amount of variability in $Y$ *before* predicting with our regression model is

$$SSTotal = \sum_i (y_i -\overline{y})^2$$
]

---

### Thinking about SSE

.large[
$$SSE = \sum_i (y_i - \widehat{y}_i)^2$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

---

### Thinking about SSTotal

.large[
$$SSTotal = \sum_i (y_i -\overline{y})^2$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ 1, lwd=1.2)
```

---

### SSE vs. SSTotal

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)

m1 <- lm(CO2 ~ poly(Day, 2, raw=TRUE), data = CO2Germany)
#sum(m1$residuals^2)
```

SSE = 4992.71
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ 1, lwd=1.2)

m2 <- lm(CO2 ~ 1, data = CO2Germany)
#sum(m2$residuals^2)
```

SSTotal: 11703.69
]

.large[
.question[
Do I want SSE to be large relative to SSTotal, or small relative to SSTotal?
]
]

---

### $R^2$: the coefficient of determination

.large[
$$SSE = \sum_i (y_i - \widehat{y}_i)^2$$
$$SSTotal = \sum_i (y_i -\overline{y})^2$$
]

.large[
The proportion of variability in $Y$ "explained" by our regression model is 

$$R^2 = 1 - \dfrac{SSE}{SSTotal}$$
]

.large[
* $R^2$ is between 0 and 1
* A higher $R^2$ means our model is a "better fit" to the data
* For regression with one variable, $R^2 = r^2$ (the squared correlation)
]

---

### SSE vs. SSTotal

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)

m1 <- lm(CO2 ~ poly(Day, 2, raw=TRUE), data = CO2Germany)
#sum(m1$residuals^2)
```

SSE = 4992.71
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ 1, lwd=1.2)

m2 <- lm(CO2 ~ 1, data = CO2Germany)
#sum(m2$residuals^2)
```

SSTotal: 11703.69
]

.large[
$R^2 = 1 - \dfrac{4992.71}{11703.69} = 0.573$
]

---

### Comparing $R^2$ values

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$R^2 = 0.573$
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
$R^2 = 0.575$
]
]

.large[
.question[
$R^2$ has increased very slightly -- does that mean that a polynomial of order 3 is a better fit?
]
]

---

### $R^2$

.large[
$R^2$ **will always increase when you add more parameters to the model.**

Why? Because

$$R^2 = 1 - \dfrac{SSE}{SSTotal}$$
and adding more parameters gives us more flexibility to decrease SSE.
]

---

### Adjusted $R^2$

.large[
We can account for the number of parameters in the model with *adjusted* $R^2$:

$$R^2_{adj} = 1 - \dfrac{SSE/(n-p)}{SSTotal/(n-1)}$$
]

.large[
* $n =$ number of observations in data (sample size)
* $p =$ number of parameters ( $\beta$ terms) in model
]

---

### Adjusted $R^2$

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
SSE = 4992.7

SSTotal = 11703.69
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
SSE = 4971.9

SSTotal = 11703.69
]
]

.large[
.question[
Why is SSTotal the same for the two models?
]
]

---


### Adjusted $R^2$

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
SSE = 4992.7

SSTotal = 11703.69
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
SSE = 4971.9

SSTotal = 11703.69
]
]

.large[
$$R^2_{adj} = 1 - \dfrac{SSE/(n-p)}{SSTotal/(n-1)}$$
]

---

### Adjusted $R^2$

.pull-left[
Polynomial of order 2:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2)
```

.large[
$R^2_{adj} = 0.56976$
]

]

.pull-right[
Polynomial of order 3:
```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red")
```

.large[
$R^2_{adj} = 0.56972$
]
]







