---
title: Inference for linear regression models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Activity

.large[
Work on the activity (handout), then we will discuss as a class.
]

---

### Activity

.large[
Your parents, worried about the nutritional value of your breakfast, believe that there will be a positive relationship between the amount of sugar in a serving, and the number of calories per serving.

.question[
What is our population model?
]
]

---

### Activity

.large[
```
Coefficients:
            Estimate Std. Error   
(Intercept)  87.4277     5.1627  
Sugar         2.4808     0.7074 
```

.question[
How would I interpret the estimated slope?
]
]

---

### Activity

.large[
Your parents, worried about the nutritional value of your breakfast, believe that there will be a positive relationship between the amount of sugar in a serving, and the number of calories per serving.

$$\text{Calories} = \beta_0 + \beta_1 \text{Sugar} + \varepsilon$$

.question[
What are our null and alternative hypotheses?
]
]

---

### Activity

.large[
$$H_0: \beta_1 = 0 \hspace{1cm} H_A: \beta_1 > 0$$

```
Coefficients:
            Estimate Std. Error   
(Intercept)  87.4277     5.1627  
Sugar         2.4808     0.7074 
```

.question[
What is our test statistic? What are the degrees of freedom?
]
]

---

### Activity

```{r, echo=F, message=F, warning=F, fig.width = 7, fig.height=4}
library(tidyverse)
x <- qt(ppoints(1500), df=34)
y <- dt(x, df=34)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  labs(x = "", y = "") +
  geom_vline(xintercept = 3.507) +
  theme(text = element_text(size = 20))
```

.large[
.question[
What area under the curve corresponds to my p-value?
]
]

---

### A new question

.large[
$$\text{Calories} = \beta_0 + \beta_1 \text{Sugar} + \varepsilon$$

```
Coefficients:
            Estimate Std. Error   
(Intercept)  87.4277     5.1627  
Sugar         2.4808     0.7074 
```

.question[
Here $\widehat{\beta}_1 = 2.48$. Will $\widehat{\beta}_1$ be exactly equal to the true value $\beta_1$?
]
]

---

### A new question

.large[
$$\text{Calories} = \beta_0 + \beta_1 \text{Sugar} + \varepsilon$$

```
Coefficients:
            Estimate Std. Error   
(Intercept)  87.4277     5.1627  
Sugar         2.4808     0.7074 
```

.question[
Here $\widehat{\beta}_1 = 2.48$. Will $\widehat{\beta}_1$ be exactly equal to the true value $\beta_1$?
]

No -- our estimate comes from one sample of data, and there is variability from sample to sample.

.question[
How can I account for this variability when reporting an estimate of $\beta_1$?
]
]

---

### Interval estimates

.large[
**Point estimate:** a single "best guess" value of a parameter. The estimated slope $\widehat{\beta}_1$ is a point estimate of $\beta_1$

**Interval estimate:** a "reasonable range" of possible values for a parameter. A *confidence interval* is an interval estimate

.question[
How would I construct a confidence interval for $\beta_1$?
]
]

---

### The critical value $t^*$

.large[
To construct a $(1 - \alpha)\cdot 100\%$ confidence interval, $t^*$ is the value such that the area in the right tail of the $t_{n-2}$ distribution is $\alpha/2$
]

.pull-left[
```{r, echo=F, message=F, warning=F, fig.width=6, fig.height=4}
library(tidyverse)
x <- qt(ppoints(1000), df=34)
y <- dt(x, df=34)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  geom_vline(xintercept = c(-2.03, 2.03), color="red") +
  labs(x = "", y = "", title = "Critical value for a 95% CI, df=34") +
  theme(text = element_text(size = 20))
```

]


.pull-right[
```{r, echo=F, fig.width=6, fig.height=4}
x <- qt(ppoints(1000), df=34)
y <- dt(x, df=34)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  geom_vline(xintercept = c(-2.73, 2.73), color="red") +
  labs(x = "", y = "", title = "Critical value for a 99% CI, df=34") +
  theme(text = element_text(size = 20))
```
]

---

### Computing $t^*$


.large[
Example: for a 95% confidence interval and 34 degrees of freedom, $t^*$ = 2.03

```{r}
qt(0.025, df=34, lower.tail=FALSE)
```

Example: for a 99% confidence interval and 34 degrees of freedom, $t^*$ = 2.73

```{r}
qt(0.005, df=34, lower.tail=FALSE)
```

]

---

### Confidence interval

.large[
```
Coefficients:
            Estimate Std. Error   
(Intercept)  87.4277     5.1627  
Sugar         2.4808     0.7074 
```

95% confidence interval for $\beta_1$:
]

---

### Confidence interval

.large[
```
Coefficients:
            Estimate Std. Error   
(Intercept)  87.4277     5.1627  
Sugar         2.4808     0.7074 
```

95% confidence interval for $\beta_1$:

$$2.48 \pm 2.03(0.71) = (1.04, 3.92)$$

.question[
How do I interpret this confidence interval?
]
]

---

### Confidence interval

.large[
```
Coefficients:
            Estimate Std. Error   
(Intercept)  87.4277     5.1627  
Sugar         2.4808     0.7074 
```

95% confidence interval for $\beta_1$:

$$2.48 \pm 2.03(0.71) = (1.04, 3.92)$$

Interpretation: We are 95% confident that a one-gram increase in sugar per serving of breakfast cereal is associated with an increase of between 1.04 and 3.92 calories per serving.

.question[
What does it mean to be "95% confident"?
]
]

---

### Activity

.large[
Work on the activity (handout), then we will discuss as a class.
]

---

### Activity

.large[
Data on 116 sparrows. Want to make a 98% confidence interval.

```{r, eval=F}
qt(..., df = ..., lower.tail=FALSE)
```

.question[
What do I fill in to calculate the critical value $t^*$?
]
]

---

### Activity

.large[
Data on 116 sparrows. Want to make a 98% confidence interval.

```{r}
qt(0.01, df = 114, lower.tail=FALSE)
```

]

---

### Activity

.large[
```
            Estimate Std. Error 
(Intercept)  8.75465    1.39601  
Weight       1.31341    0.09756 
```

$t^* = 2.36$

.question[
What is my 98% confidence interval?
]
]

---

### Activity

.large[
```
            Estimate Std. Error 
(Intercept)  8.75465    1.39601  
Weight       1.31341    0.09756 
```

$t^* = 2.36$

$$1.31 \pm 2.36(0.098) = (1.08, 1.54)$$

.question[
Your friend claims that there is a 2% chance that the true slope $\beta_1$ is greater than 1.54 or less than 1.08. Is your friend correct?
]

]





