---
title: Fitting logistic regression models
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Logistics and reminders

.large[
* **Optional** HW 6 due on Friday (4/24)

* Project Part 4 due Wednesday 4/29 (LDOC)

* Project work time in class after we finish up logistic regression
]

---

### Last time: Class activity

.large[
| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 3 | 2 | 3 | 1 | 1 |
| Number of failures | 0 | 0 | 2 | 1 | 2 |

What's the probability of the observed data, if $\beta_0 = 3$ and $\beta_1 = -0.5$?
]

---

### Class activity

.large[
| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 3 | 2 | 3 | 1 | 1 |
| Number of failures | 0 | 0 | 2 | 1 | 2 |

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| $\pi$ | $\dfrac{\exp\{3 - 0.5(3)\}}{1 + \exp\{3 - 0.5(3)\}} = 0.818$ | 0.731 | 0.622 | 0.5 | 0.378 |

$L(\beta_0, \beta_1) = P(\text{data}) =$
$$(0.818)^3(0.731)^2(0.622)^3(1 - 0.622)^2(0.5)(1 - 0.5) \\ \cdot (0.378)(1 - 0.378)^2$$
$= 0.000368$
]

---

### Class activity

.large[
| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 3 | 2 | 3 | 1 | 1 |
| Number of failures | 0 | 0 | 2 | 1 | 2 |

What's the probability of the observed data, if $\beta_0 = 4$ and $\beta_1 = -0.75$?
]

---

### Class activity

.large[
| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 3 | 2 | 3 | 1 | 1 |
| Number of failures | 0 | 0 | 2 | 1 | 2 |

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| $\pi$ | $\dfrac{\exp\{4 - 0.75(3)\}}{1 + \exp\{4 - 0.75(3)\}} = 0.852$ | 0.731 | 0.562 | 0.378 | 0.223 |

$L(\beta_0, \beta_1) = P(\text{data}) =$
$$(0.852)^3(0.731)^2(0.562)^3(1 - 0.562)^2(0.378)(1 - 0.378) \\ \cdot (0.223)(1 - 0.223)^2$$
$= 0.000356$
]

---

### Maximum likelihood for logistic regression

.large[
**Likelihood:** 
* For estimates $\widehat{\beta}_0$ and $\widehat{\beta}_1$, $\widehat{\pi} = \dfrac{\exp\{\widehat{\beta}_0 + \widehat{\beta}_1 x\}}{1 + \exp\{\widehat{\beta}_0 + \widehat{\beta}_1 x\}}$
* $L(\widehat{\beta}_0, \widehat{\beta}_1) = P(\text{data})$
    
**Maximize:** 
* Choose $\widehat{\beta}_0$, $\widehat{\beta}_1$ to maximize $L(\widehat{\beta}_0, \widehat{\beta}_1)$
    
.question[
In the class activity examples, we have only considered a few different possibilities for $\widehat{\beta}_0$ and $\widehat{\beta}_1$. In reality, we consider all real numbers as possible values.
]
]

---

### Fitting logistic regression

.large[
.pull-left[
**Logistic regression with maximum likelihood**

* **Model:** $\log \left( \dfrac{\pi}{1 - \pi} \right) = \beta_0 + \beta_1 x$
* **Measure of fit:** $L(\widehat{\beta}_0, \widehat{\beta}_1) =$ probability of observed data for estimates $\widehat{\beta}_0$, $\widehat{\beta}_1$
* **Optimize:** choose $\widehat{\beta}_0$, $\widehat{\beta}_1$ to maximize $L(\widehat{\beta}_0, \widehat{\beta}_1)$
]

.pull-right[
**Linear regression with least squares**

<br/>

* **Model:** $y = \beta_0 + \beta_1 x + \varepsilon$
* **Measure of fit:** $SSE = \sum \limits_{i = 1}^n (y_i - \widehat{\beta}_0 - \widehat{\beta}_1 x_i)^2$
* **Optimize:** choose $\widehat{\beta}_0$, $\widehat{\beta}_1$ to minimize SSE
]
]

---

### Regression assumptions

.large[
.question[
What assumptions did we make for linear regression models?
]
]

---

### Regression assumptions

.large[
.question[
What assumptions did we make for linear regression models?
]
]

.large[
* Shape
* Constant variance
* Normality
* Independence
* Randomness

.question[
Not all these assumptions translate to logistic regression.
]
]

---

### Assumptions: logistic vs. linear regression

.large[
.pull-left[
**Logistic regression:**

$\pi = P(y = 1)$ 

$\log \left( \dfrac{\pi}{1 - \pi} \right) = \beta_0 + \beta_1 x$

* shape
* randomness
* independence
]

.pull-right[
**Linear regression:**

$y = \beta_0 + \beta_1 x + \varepsilon$

$\varepsilon \sim N(0, \sigma_\varepsilon^2)$

* shape
* randomness
* independence
* constant variance
* normality

]

.question[
The constant variance and normality assumptions apply specifically to the noise term $\varepsilon$
]
]

---

### Data

.large[
Relationship between the length of a putt, and whether it was made:

* $x =$ length (feet)
* $y =$ result (success = 1, failure = 0)

$\log \left( \dfrac{\pi}{1 - \pi} \right) = \beta_0 + \beta_1 x$

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
]

---

### Assessing the shape assumption

.large[
.question[
How did we assess the shape assumption for linear regression?
]
]

---

### Assessing the shape assumption

.large[
.question[
How did we assess the shape assumption for linear regression?
]
]

.large[
With scatterplots and residual plots.

.question[
Can we use these diagnostics for logistic regression?
]
]

---

### Assessing the shape assumption

.large[
.question[
How did we assess the shape assumption for linear regression?
]
]

.large[
With scatterplots and residual plots.

.question[
Can we use these diagnostics for logistic regression?
]
]

.large[
No: 
* the response is binary
* the shape assumption is that the *log odds* (not the response) are a linear function of the predictor(s)
]

---

### Assessing the shape assumption

.large[
Scatterplots don't work for logistic regression:

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

library(tidyverse)
library(Stat2Data)
data("Putts1")

Putts1 %>%
  ggplot(aes(x = Length, y = Made)) +
  geom_point(size = 2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

]

---

### Shape

.large[
* Shape assumption is that the log-odds are (at least approximately) a linear function of $x$: $\log(odds) \approx \beta_0 + \beta_1 x$
* We can assess the shape assumption by plotting the **empirical log-odds** (aka empirical logits)
* Example: relationship between the length of a putt and whether it was made

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
]

---

### Empirical log-odds

.large[
**Step 1:** estimate the probability of success for each length of putt

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
| Probability of success $\widehat{p}$ | 0.832 | 0.739 | 0.565 | 0.488 | 0.328 |
]

---

### Empirical log-odds

.large[
**Step 2:** convert empirical probabilities to empirical odds

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
| Probability of success $\widehat{p}$ | 0.832 | 0.739 | 0.565 | 0.488 | 0.328 |
| Odds $\dfrac{\widehat{p}}{1 - \widehat{p}}$ | 4.941 | 2.839 | 1.298 | 0.953 | 0.489 |
]

---

### Empirical log-odds

.large[
**Step 3:** calculate empirical log-odds

| Length | 3 | 4 | 5 | 6 | 7 |
| --- | --- | --- | --- | --- | --- |
| Number of successes | 84 | 88 | 61 | 61 | 44 |
| Number of failures | 17 | 31 | 47 | 64 | 90 |
| Total | 101 | 119 | 108 | 125 | 134 |
| Probability of success $\widehat{p}$ | 0.832 | 0.739 | 0.565 | 0.488 | 0.328 |
| Odds $\dfrac{\widehat{p}}{1 - \widehat{p}}$ | 4.941 | 2.839 | 1.298 | 0.953 | 0.489 |
| Log-odds $\log \left( \dfrac{\widehat{p}}{1 - \widehat{p}} \right)$ | 1.60 | 1.04 | 0.26 | -0.05 | -0.72 |
]

---

### Empirical log-odds

.large[
**Step 4:** plot empirical log-odds against predictor

```{r echo=F, message=F, fig.width=7, fig.height=5, fig.align='center'}
data.frame(length = c(3, 4, 5, 6, 7),
  log_odds = c(1.6, 1.04, 0.26, -0.05, -0.72)) %>% 
  ggplot(aes(x = length, y = log_odds)) + 
  geom_point(size=3) + 
  geom_smooth(se=F, method="lm") +
  labs(x = "Length (feet)",
       y = "Empirical log-odds") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.question[
The shape assumption is reasonable if the empirical log-odds plot looks roughly linear.
]
]

---

### Empirical log odds plots

.large[
Now let's try this with the med school admissions data:

| GPA | 2.26 | 2.42 | 2.48 | 2.52 | 2.55 | 2.56 | ... |
| --- | --- | --- | --- | --- | --- | --- | --- |
| Admit = 0 | 1 | 1 | 1 | 1 | 1 | 1 | ... |
| Admit = 1 | 0 | 1 | 0 | 0 | 0 | 0 | ... |

.question[
What problem do I run into?
]
]

---

### Empirical log odds plots

.large[
Now let's try this with the med school admissions data:

| GPA | 2.26 | 2.42 | 2.48 | 2.52 | 2.55 | 2.56 | ... |
| --- | --- | --- | --- | --- | --- | --- | --- |
| Admit = 0 | 1 | 1 | 1 | 1 | 1 | 1 | ... |
| Admit = 1 | 0 | 1 | 0 | 0 | 0 | 0 | ... |

.question[
What problem do I run into?
]
]

.large[
There are a small number of observations for each GPA, so we can't calculate empirical log odds for each GPA.

**Solution:** Divide the data into *bins*
]

---

### Empirical log odds plots

.large[
* Divide observations into bins of equal size, based on values of predictor $X$
* Calculate empirical log odds in each bin
* Plot empirical log odds

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
num_bins <- 8

grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

logodds_table <- grad_app %>%
  mutate(obs = admit,
         bin = cut(gpa, 
                   breaks = quantile(gpa, 
                                     seq(0, 1,
                                         length.out=(num_bins+1))),
                   labels = F, include.lowest=T)) %>%
  group_by(bin) %>%
  summarize(mean_x = mean(gpa),
            prop = mean(c(obs, 0.5)),
            num_obs = n()) %>%
  ungroup() %>%
  mutate(logodds = log(prop/(1 - prop)))

logodds_table %>%
  ggplot(aes(x = mean_x,
             y = logodds)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm") +
  theme_bw() +
  labs(x = "GPA",
       y = "Empirical log odds",
       color = "",) +
  theme(text = element_text(size=20))
```
]