---
title: "Activity: Exploring penguins data"
format: html
editor: source
---

**Instructions:** 

* Work with a neighbor to answer the following questions 
* Unlike other class activities, you do **not** need to submit this one

This activity will introduce you to working with data in RStudio. We'll explore some data on penguins in Antarctica. Work through the questions below, and then we'll discuss together as a group. If you get stuck, some questions have a solution available, which you can view by clicking the `Solution` button.

## Downloading the data

There are two main ways we will access data in R in this course:

* Through a package which loads it directly

* By importing a data file (e.g., a .csv or .txt file)

**Packages:** Many features of R are available through modular collections of code called *packages*. Packages can contain functions, documentation, and datasets. Packages are downloaded to your computer with `install.packages("package_name")`, and loaded into your R session with `library(package_name)`.

For this activity, the data is available through the `palmerpenguins` package.

1. In the console, run the following code. Remember: to run code in the console, type it in and hit enter, or copy and paste and hit enter.

```{r, eval=F}
install.packages("palmerpenguins")
library(palmerpenguins)
```

2. The dataset loaded by palmerpenguins is called `penguins`. In the console, run

```{r, eval=F}
?penguins
```

How many rows are in the data? What does each row represent?

<SCRIPT>
function ShowAndHide(divid) {
    var x = document.getElementById(divid);
    if (x.style.display == 'none') {
        x.style.display = 'block';
    } else {
        x.style.display = 'none';
    }
}
</SCRIPT>

<BUTTON ONCLICK="ShowAndHide('q2_solution')">Solution</BUTTON>

<DIV ID="q2_solution" STYLE="display:none">
There are 344 rows in the data. Each row represents one penguin observed near Palmer Station, Antarctica. The different columns contain different measurements (species, body mass, etc.) 
</DIV>


3. Now let's look at our data. To do so, we'll need to install the `tidyverse` package, which has tools for data wrangling and visualization that we will use throughout the semester. By adapting the code from Exercise 1, install and load the `tidyverse` package. Then run 

```{r, eval=F}
glimpse(penguins)
```

What information about the data does this provide?

<BUTTON ONCLICK="ShowAndHide('q3_solution')">Solution</BUTTON>

<DIV ID="q3_solution" STYLE="display:none">
`glimpse` shows the number of rows and columns in the data, each column name, and the first few entries for each column. 
</DIV>


4. It is hard to learn a lot about data from just the data table. Instead, let's make a plot! We'll start by looking at the distribution of bill length, which is a quantitative numeric variable, so we can look at the distribution with a histogram. Run

```{r, eval=F}
penguins |>
  ggplot(aes(x = bill_length_mm)) +
  geom_histogram()
```

Describe the distribution of bill length in the penguins.

**Code explanation:** What is this code doing? The `|>`, called the *pipe*, means "take this, THEN do that". So, `penguins |> ggplot(...)` means "take penguins, THEN make a plot". The `ggplot()` function provides lots of plot options. To make a histogram, we use `geom_histogram()`. We will cover more data wrangling and ggplot in Week 2.

5. We can also look at the relationships between variables. We'll look at bill length and bill depth. Run

```{r, eval=F}
penguins |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()
```

Describe the relationship between bill length and bill depth.

6. From Exercises 4 and 5, it looks like there could be different groups in the data. Perhaps different species of penguin?
Fill in the `...` in the following code so that the color and shape depend on species:

```{r, eval=F}
penguins |>
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm, 
             color = ...,
             shape = ...)) +
  geom_point()
```


*Hint*: remember that you can find information on the variables in the dataset by running `?penguins`.

<BUTTON ONCLICK="ShowAndHide('q6_solution')">Solution</BUTTON>

<DIV ID="q6_solution" STYLE="display:none">
```{r, eval=F}
penguins |>
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm, 
             color = species,
             shape = species)) +
  geom_point()
```
</DIV>

7. Now suppose that I give you a penguin with bill length 50mm and bill depth 15mm. What species would you guess it is? (This is an example of a *prediction* question, one important use for statistical models.)

**If you're finished early:** try to customize the plots! Search online to see if you can figure out how to change:

* the axis labels
* the theme
* the colors of the points

Hint 1: googling "ggplot plot labels" is a good place to start. Similar searches would be useful for the theme and colors.
