---
title: "Class activity: inference for the regression slope"
format: pdf
editor: source
---

\vspace{-2.5cm}

**Group members:**

**Instructions:** Work with a neighbor on the following activity. I will collect the handout at the end of class, and it will be part of your class participation grade. You will be graded only on effort -- it is ok if you don't finish all the questions, or get them all correct.

## Sparrow data

In this activity, you will practice hypothesis testing with a linear regression model. We will work with data on 116 sparrows, which were observed on Kent Island, New Brunswick. For each sparrow, we have information including:

* *Weight*: the weight of the sparrow (in grams)
* *WingLength*: the sparrow's wing length (in mm)

We fit the model

$$\text{wing length} = \beta_0 + \beta_1 \text{weight} + \varepsilon$$
Here is output from the fitted model:

```{r, include=F}
library(Stat2Data)
library(tidyverse)
data("Sparrows")

sparrow_lm <- lm(WingLength ~ Weight, data = Sparrows)
summary(sparrow_lm)$coefficients
```

\begin{verbatim}
Coefficients:
            Estimate Std. Error  
(Intercept)  8.75465    1.39601  
Weight       1.31341    0.09756 
\end{verbatim}


## Questions

Researchers want to know whether there is a relationship between sparrow weight and wing length.

1. Write down null and alternative hypotheses corresponding to the research question.

\vspace{3cm}

2. Using the regression output, calculate the test statistic for the hypotheses in question 1.

\vspace{3cm}

3. Without calculating a p-value, do you think the observed test statistic would be unusual if the null hypothesis were true?

\vspace{2cm}

4. What will be the degrees of freedom for the $t$ distribution used to calculate a p-value in this test?

\vspace{1.5cm}

5. Below is a picture of the appropriate $t$ distribution. Indicate the area under the curve corresponding to the p-value for your hypothesis test.

```{r, echo=F, message=F, warning=F}
library(tidyverse)
x <- seq(-15, 15, by=0.05)
y <- dt(x, df=114)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  labs(x = "", y = "")
```



