---
title: "Class activity: Exploring confidence intervals"
format: html
editor: source
---

\vspace{-2.5cm}

**Instructions:** 

* Work individually or with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta112-s26.github.io/class_activities/ca_27_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

# First Year GPA 

Previously, we have worked with data on the first year GPAs for a sample of 219 college students. The `FirstYearGPA` dataset, in the `Stat2Data` package, includes the variables:

* `GPA`: the student's GPA in their first year of college
* `HSGPA`: the student's high school GPA
* `SATV`: the student's verbal SAT score

Here are two scatterplots showing the relationships between these variables

```{r, message=F, warning=F, fig.width=9, fig.height=4}
library(tidyverse)
library(patchwork)
library(Stat2Data)
data("FirstYearGPA")

p1 <- FirstYearGPA |>
  ggplot(aes(x = HSGPA, y = GPA)) +
  geom_point() +
  theme_bw()

p2 <- FirstYearGPA |>
  ggplot(aes(x = SATV, y = GPA)) +
  geom_point() +
  theme_bw()

p1 + p2
```

We are interested in predicting first year GPA from high school GPA and verbal SAT score, so we fit the model

$$\text{GPA} = \beta_0 + \beta_1 \text{HSGPA} + \beta_2 \text{SATV} + \varepsilon$$

1. **Question 1:** Fit the linear model in R, using the `FirstYearGPA` data.

We now wish to examine the relationship between high school GPA and first-year college GPA, after accounting for SAT score (to potentially account for different grading systems in different high schools). 

2. **Question 2:** Create and interpret a 95% confidence interval for the average change in first-year GPA associated with a 1-unit increase in high school GPA, holding verbal SAT score fixed. You can get the critical value $t^*$ with `qt(..., df = ..., lower.tail=F)`, where you will fill in the `qt` function with the desired amount in each tail and the degrees of freedom for the relevant $t$-distribution.

# Understanding confidence intervals

You may wonder why we use a $t$-distribution, or why our confidence interval is $\widehat{\beta} \pm t^* SE_{\widehat{\beta}}$, or where we get $SE_{\widehat{\beta}}$ from in the first place. To try to explore those questions, the next part of the class activity will do some simulation.

We can think of a confidence interval as providing **reasonable range** of values for an unknown parameter. The parameter isn't guaranteed to be included in this range, but this is the range where we think it *might* be. We use a confidence interval because, while we calculate an estimate $\widehat{\beta}$ from our sample data, we know that estimate will change if we were to take another sample.

## Bootstrapping

In actual practice, we don't get to take another sample from the population -- we only have the one sample of data. **But**, we can try to approximate the process of sampling from the population, by sampling *with replacement* from our data. This is called **bootstrapping**. 

The code below creates a bootstrap sample, then re-makes the two scatterplots:

```{r, message=F, warning=F, fig.width=9, fig.height=4}
# create the bootstrap data
bootstrap_data <- FirstYearGPA |>
    sample_n(n(), replace = TRUE)

# plot the bootstrap data
p1 <- bootstrap_data |>
  ggplot(aes(x = HSGPA, y = GPA)) +
  geom_point() +
  theme_bw()

p2 <- bootstrap_data |>
  ggplot(aes(x = SATV, y = GPA)) +
  geom_point() +
  theme_bw()

p1 + p2
```

These plots look similar to the ones above, but not identical, because we have slightly different data.

3. **Question 3:** Run the code to create the bootstrap data, then re-fit the linear model from question 1 using the bootstrap data. Explain why the estimated regression coefficients are similar, but not identical.

The key idea of bootstrapping is that we can approximate the variability of a statistic, like $\widehat{\beta}$, by looking at the variability in many bootstrap samples. The following code produces 10000 bootstrap samples, and computes $\widehat{\beta}_1$ for each sample. We then compute the standard deviation of these bootstrap estimates $\widehat{\beta}_1$:

```{r, eval=F}
nsim <- 10000
bootstrap_coefficients <- rep(NA, nsim)
for(i in 1:nsim){
  bootstrap_data <- FirstYearGPA |>
    sample_n(n(), replace = TRUE)
  m_boot <- lm(GPA ~ HSGPA + SATV, data = bootstrap_data)
  bootstrap_coefficients[i] <- coef(m_boot)[2]
}

# standard deviation of bootstrap estimates
sd(bootstrap_coefficients)
```

4. **Question 4:** Run the code (it may take a few seconds; if it takes too long, decrease the number of simulations). How does the standard deviation compare with the standard error $SE_{\widehat{\beta}_1}$ you used to compute the confidence interval in question 2?

Now let's look at the distribution of the bootstrap coefficients:

```{r, eval=F}
hist(bootstrap_coefficients)
```


5. **Question 5:** What does the distribution look like?

## Bootstrap percentile interval

Here is an intuitive way to create a 95% confidence interval for $\beta_1$: make the lower endpoint the 2.5th percentile of the bootstrap samples, and the upper endpoint the 97.5th percentile. That is: the confidence interval is just the middle 95% of bootstrap samples! If the bootstrap is trying to approximate sampling from the population, then the middle of the bootstrap distribution should be "reasonable range" of values for $\beta_1$.

Here is code to compute those endpoints:

```{r, eval=F}
quantile(bootstrap_coefficients, 0.025)
quantile(bootstrap_coefficients, 0.975)
```


6. **Question 6:** How do the 2.5th percentile and 97.5th percentile compare to your confidence interval endpoints from question 2?



