---
title: "Class activity: Exploring nested F tests"
format: html
editor: source
---

\vspace{-2.5cm}

**Instructions:** 

* Work individually or with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta112-s26.github.io/class_activities/ca_26_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

# Testing with the penguins data

Suppose we are working with the penguins data, and we fit the model

$$\text{body mass} = \beta_0 + \beta_1 \text{flipper length} + \beta_2 \text{IsChinstrap} + \beta_3 \text{IsGentoo} + \varepsilon$$

The plot below shows the relationship between these variables in the penguins data:

```{r, message=F, warning=F}
library(palmerpenguins)
library(tidyverse)
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  labs(x = "Flipper length", y = "Body mass") +
  theme_bw()
```

We wisht to test the hypotheses

$$H_0: \beta_2 = \beta_3 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_2, \beta_3 \neq 0$$

This means we are comparing the full and reduced models:

$$\text{Full model:} \hspace{1cm} \text{body mass} = \beta_0 + \beta_1 \text{flipper length} + \beta_2 \text{IsChinstrap} + \beta_3 \text{IsGentoo} + \varepsilon$$

$$\text{Reduced model:} \hspace{1cm} \text{body mass} = \beta_0 + \beta_1 \text{flipper length} + \varepsilon$$

In R, we can fit these full and reduced models:

```{r}
m_full <- lm(body_mass_g ~ flipper_length_mm + species, data = penguins)
m_reduced <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
```

## Comparing the models

Now, how should we compare the full and reduced models? Intuitively, we want to look at which model is a *better fit to the data*. To measure model fit, we could use SSE:

$$SSE_{reduced} = \sum_i(y_i - \widehat{y}_{i,reduced})^2 \hspace{1cm} SSE_{full} = \sum_i(y_i - \widehat{y}_{i,full})^2$$

We want SSE to be small; this suggests we could use the *reduction* in SSE for the full model vs. the reduced model as a way to compare the two models:

$$SSE_{reduced} - SSE_{full}$$

However, we also know that SSE will decrease when we add more variables to the model. So:

* If the SSE only goes down by a *small* amount, then maybe the full model is not much better than the reduced model, and we should just pick the reduced model
* If the SSE goes down by a *big* amount, then the full model does a substantially better job than the reduced model

The code below calculates the SSE for both models on the penguins data:

```{r}
# calculate the SSE
sse_red <- sum(m_reduced$residuals^2)
sse_full <- sum(m_full$residuals^2)
sse_red
sse_full
```

1. **Question 1:** What is the change in SSE between the two models? Do you think this is a big or small change in SSE?

## F statistic

**However:** remember that SSE depends on the number of observations in the data, *and* the scale of the response variable! In order to understand what a "big" or "small" change is, we can compute a modified version of the change in SSE, which is our **F statistic**:

$$F = \dfrac{(SSE_{reduced} - SSE_{full})/(p_{full} - p_{reduced})}{SSE_{full}/(n - p_{full})}$$

where $n$ is the number of observations in the data, $p_{full}$ is the number of parameters ($\beta$s) in the full model, and $p_{reduced}$ is the number of parameters in the reduced model. By incorporating information about degrees of freedom and $SSE_{full}$ in the denominator, we are accounting for how the difference in complexity between the two models, the number of observations, and the scale of the response variable.

2. **Question 2:** Use `sse_red` and `sse_full` to compute the F statistic. Here $n = 342$.

3. **Question 3:** Now use the `anova` function (code below) to verify the F statistic you calculated in question 2, and get the p-value. Would you reject or fail to reject the null hypothesis?

```{r, eval=F}
anova(m_reduced, m_full)
```

## Permuting the data

We have some sense of where the F statistic comes from, but to calculate a p-value we need to know its distribution! In particular, we need to know how $F$ behaves when $H_0$ is true.

So, how do we know the distribution of $F$ when $H_0$ is true? We have two options:

* Do some fairly involved math (probably not at the 112 level!)
* *Make* $H_0$ true and see what happens (this we can do in R!)

To make $H_0$ true, we will use a technique called **permutation**: we will create a new version of the penguins dataset, in which the species of each row is *randomly* assigned. Because species is randomly assigned in our permuted dataset, there can't be any relationship between species and body mass!

The code below does the permutation, and shows a plot of the permuted data:

```{r, warning=F, message=F}
penguins_permuted <- penguins |>
  mutate(species = sample(species, n(), replace=F))

penguins_permuted |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  labs(x = "Flipper length", y = "Body mass") +
  theme_bw()
```

We can also re-fit the full and reduced models with the permuted data, and compare them:

```{r, eval=F}
m_full <- lm(body_mass_g ~ flipper_length_mm + species, data = penguins_permuted)
m_reduced <- lm(body_mass_g ~ flipper_length_mm, data = penguins_permuted)
sum(m_reduced$residuals^2) - sum(m_full$residuals^2) # difference in SSE
anova(m_reduced, m_full) # F statistic and p-value
```

4. **Question 4:** How does the difference in SSE compare to your value in question 1? How does the F statistic compare to the value in question 2? Explain why this makes sense.

## Repeating many times

To decide whether to reject $H_0$, we need to decide whether the $F$ statistic from question 2 is sufficiently *unusual*, under the null hypothesis. However, the F statistic you calculated in question 4 is just one possible value we could have observed, *if* the null hypothesis were true. To really decide whether a test statistic is unusual, we need to repeat the process many times, and get *many* statistics under $H_0$! 

The code below allows you to do that efficiently:

```{r, eval=F}
nsim <- 1000
f_stats <- rep(NA, nsim)
for(i in 1:nsim){
  penguins_permuted <- penguins |>
  mutate(species = sample(species, n(), replace=F))
  
  m_full <- lm(body_mass_g ~ flipper_length_mm + species, data = penguins_permuted)
  m_reduced <- lm(body_mass_g ~ flipper_length_mm, data = penguins_permuted)
  f_stats[i] <- anova(m_reduced, m_full)$F[2]
}

hist(f_stats, xlab = "F", main = "Distribution of F statistic under null H0")
```

5. **Question 5:** Run the code, producing a histogram of 1000 different $F$ statistics under the null hypothesis. From this distribution, does the $F$ statistic from question 2 seem *unusual* under the null hypothesis? Explain.

## The F distribution

The permutation method we have used here is a great way to look at the behavior of a test statistic under $H_0$. However, it does require us to do a lot of computation. Sometimes, having a mathematical description of the distribution is helpful.

For the $F$ test, it turns out that, if $H_0$ is true, then $F \sim F_{p_{full} - p_{reduced}, n - p_{full}}$. The code below samples 1000 random draws from this distribution, and makes a histogram:

```{r, eval=F}
# example sample from F distribution
f_samp <- rf(1000, 2, 338)
hist(f_samp)
```

6. **Question 6:** Run the code, and verify that the distribution looks similar to the distribution from question 5.
