---
title: "Class activity: logistic regression diagnostics"
format: html
editor: source
---

**Instructions:** 

* Work individually or with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta112-s26.github.io/class_activities/ca_34_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

# Titanic data

The `titanic` data, discussed in class, contains information on 891 passengers from the Titanic disaster. For each passenger, we have information that includes:

* `Survived`: whether the passenger survived (1) or not (0)
* `Sex`: passenger sex (`male` or `female`)
* `Age`: passenger age (in years)
* `Fare`: how much the passenger paid for their ticket

The `titanic` data can be loaded into `R` with the following:

```{r, eval=F}
titanic <- read.csv("https://sta112-s26.github.io/class_activities/Titanic.csv")
```

## Questions

The code below will make an empirical logit plot with Fare as the explanatory variable, Survived as the binary response, and 20 bins:

```{r, eval=F}
library(tidyverse)
logodds_plot <- function(data, reg_formula, group,
                         num_bins = 5){
  
  var_names <- all.vars(reg_formula)

  if(length(var_names) != 2){
    stop("Only one variable can appear on the x-axis")
  }

  if(!is.numeric(data[[var_names[2]]])){
    stop("The explanatory variable must be numeric")
  }
  
  if(n_distinct(data[[var_names[1]]]) != 2){
    stop("The response variable must be binary")
  }
  
  data <- data |>
    select(all_of(var_names), {{group}}) |>
    drop_na()

  dat <- model.frame(reg_formula, data)
  

  xname <- colnames(dat)[2]
  yname <- colnames(dat)[1]
  colnames(dat) <- c("y", "x")

  if(any(is.infinite(dat$x))){
    stop(paste("There are non-finite values in", xname))
  }

  if(!missing(group)){
    dat <- dat |>
      mutate({{group}} := as.factor(data |> pull({{group}})))
  }

  dat |>
    arrange({{group}}, x) |>
    group_by({{group}}) |>
    mutate(bin = rep(1:num_bins,
                     each=ceiling(n()/num_bins))[1:n()]) |>
    group_by(bin, {{group}}) |>
    summarize(x = mean(x),
                prop = (sum(y) + 0.5)/(n() + 1),
                num_obs = n(),
                .groups = "drop") |>
    mutate(logodds = log(prop/(1 - prop))) |>
    ggplot(aes(x = x,
               y = logodds,
               color = {{group}})) +
    geom_point() +
    geom_smooth(se=F, method="lm", formula = y ~ x) +
    theme_bw() +
    labs(x = xname,
         y = "Empirical log odds")
}

logodds_plot(titanic, Survived ~ Fare, 
             num_bins = 20)
```

1. Run the code to produce the empirical logit plot. Does it look like there is a linear relationship between Fare and the log odds of survival? Experiment with the number of bins in the final line of the code to be sure.

2. You can try different transformations of Fare in the empirical logit plot by changing the formula. For example, you could try `Survived ~ sqrt(Fare)`. Experiment with a few different transformations -- which ones seem reasonable?

### Passenger age and sex

3. Now make an empirical logit plot with `Age` as the explanatory variable, and `Survived` as the response. Does it look like there is a linear relationship between Age and the log odds of survival?

Sometimes, we may want to account for another variable. In the code above, you can do that with the `group` argument. For example, 

```{r, eval=F}
logodds_plot(titanic, Survived ~ Age, group = Sex,
             num_bins = 20)
```


4. Use the empirical logit plot to describe the relationship between Age, Sex, and the log odds of survival. What trends do you notice? Is this what you would expect?

5. Fit a logistic regression model with `Survived` as the response and `Age` and `Sex` as the explanatory variables. Use your fitted model to predict the probability of survival for a female passenger who is 40 years old.

