---
title: "Class Activity"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

**Instructions:** 

* Work individually or with a neighbor to answer the following questions 
* To get started, download the [class activity template](https://sta112-s26.github.io/class_activities/ca_35_template.qmd) file
* When you are finished, render the file as an HTML and submit the HTML to Canvas (let me know if you encounter any problems)

## Pitching data

In this activity, you will look at a dataset of 3402 pitches thrown by MLB pitcher Clayton Kershaw in the 2013 season. The data is contained in the `Kershaw` data set, in the `Stat2Data` R package. We will focus on two specific variables for each pitch:

* `Result`: a negative result (a ball or a hit), or a positive result (a strike or an out)
* `EndSpeed`: the speed at which the ball crossed home plate (in mph)

To load the data, run

```r
library(Stat2Data)
data("Kershaw")
```

### Questions

We are interested in the logistic regression model for the relationship between end speed and result:

$\log \left( \dfrac{\pi}{1 - \pi} \right) = \beta_0 + \beta_1 \ \text{speed}$,

where $\pi$ is the probability of a positive result.


1. You are interested in testing whether there is any relationship between whether a pitch is successful and the speed at which it crosses home plate. Carry out both a Wald test and a likelihood ratio test to investigate this question, and report your results. *In practice, we would just do one test, rather than both of them. But since this is a class activity, I want you to try both tests.*

<SCRIPT>
function ShowAndHide(divid) {
    var x = document.getElementById(divid);
    if (x.style.display == 'none') {
        x.style.display = 'block';
    } else {
        x.style.display = 'none';
    }
}
</SCRIPT>

<BUTTON ONCLICK="ShowAndHide('q1_solution')">Solution</BUTTON>

<DIV ID="q1_solution" STYLE="display:none">
**Wald test:**

```{r, include=F}
library(Stat2Data)
data("Kershaw")
```

```{r}
kershaw_glm <- glm(Result ~ EndSpeed, data = Kershaw, family=binomial)
summary(kershaw_glm)
```

test statistic: $z = 3.881$. p-value = 0.000104. This p-value is small, so we have strong evidence that there is a relationship between pitch speed and result.

**Likelihood ratio test:**

full model: $\log \left( \dfrac{\pi}{1 - \pi} \right) = \beta_0 + \beta_1 \ \text{speed}$

reduced model: $\log \left( \dfrac{\pi}{1 - \pi} \right) = \beta_0$

test statistic: $G = 4540.4 - 4525.4 = 15.0$

degrees of freedom: 1 (we are testing a single parameter), so $G \sim \chi^2_1$

p-value:

```{r}
pchisq(15, df=1, lower.tail=F)
```

</DIV>

2. Instead of the research question in exercise 1, you want to know whether faster pitches are more likely to be successful. Write down your hypotheses in terms of a parameter in the model.

3. Which test (Wald, likelihood ratio, or both) can you use to test the hypotheses in question 2? Carry out the hypothesis test.

