---
title: "Exam 2 review questions"
format: html
---

Below are questions to help you study for Exam 2. These are some examples of the kinds of questions I might ask.

* This is not a practice exam.
* The questions cover most, but not all, possible material for the exam.
* The distribution of questions here is not necessarily reflective of the distribution of questions on the actual exam.


## Multiple choice questions

1. Which of the following is the correct formula for $R^2$?
    (a) $R^2 = \dfrac{\sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{\sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (b) $R^2 = 1 - \dfrac{\sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{\sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (c) $R^2 = \dfrac{\frac{1}{n-p}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-1} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (d) $R^2 = 1 - \dfrac{\frac{1}{n-p}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-1} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (e) $R^2 = \dfrac{\frac{1}{p-1}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-1} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (f) $R^2 = 1 - \dfrac{\frac{1}{p-1}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-p} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    
2. Which of the following is the correct formula for $R_{adj}^2$?
    (a) $R_{adj}^2 = \dfrac{\sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{\sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (b) $R_{adj}^2 = 1 - \dfrac{\sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{\sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (c) $R_{adj}^2 = \dfrac{\frac{1}{n-p}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-1} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (d) $R_{adj}^2 = 1 - \dfrac{\frac{1}{n-p}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-1} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (e) $R_{adj}^2 = \dfrac{\frac{1}{p-1}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-1} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    (f) $R_{adj}^2 = 1 - \dfrac{\frac{1}{p-1}  \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2}{ \frac{1}{n-p} \sum \limits_{i=1}^n (y_i - \overline{y})^2}$
    
3. How are the $F$ and $t$ distributions related?
    (a) If $t \sim t_d$, then $t^2 \sim F_{d, 1}$
    (b) If $t \sim t_d$, then $t^2 \sim F_{1, d}$
    (c) If $F \sim F_{1,d}$ then $F^2 \sim t_d$
    (d) If $F \sim F_{d, 1}$ then $F^2 \sim t_d$
    
## Short answer questions

4. Can $R^2$ decrease when additional variables are added to a model? Explain. What about $R^2_{adj}$?

5. How should we handle potential outliers with a studentized residual $< -3$ or $> 3$? 

6. A multiple regression model has response $Y$ and two quantitative explanatory variables, $X_1$ and $X_2$:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \varepsilon$$
The sample correlation between $X_1$ and $X_2$ is $r = 0.6$. Calculate the variance inflation factors (VIFs) for $X_1$ and $X_2$.

7. A multiple regression model has response $Y$ and three quantitative explanatory variables, $X_1$, $X_2$, and $X_3$:

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \varepsilon$$

The variance inflation factor (VIF) for $X_1$ is 4. Calculate the $R^2$ for the regression of $X_1$ on $X_2$ and $X_3$.

8. What is Simpson's paradox?

9. What are some benefits of including multiple explanatory variables in a regression model?

10. True or False: any hypotheses which can be tested with a nested F-test (comparing full and reduced models) can also be tested with a t-test. Explain your answer.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

set.seed(3)

x = rnorm(100, mean=50, sd=15)
crabs <- data.frame(Noise = rep(c("ocean", "ship"), each = 50),
                       Mass = x,
                       Oxygen = c(50 + 2*x[1:50] + rnorm(50, sd=15),
                                  300 - 2*x[51:100] + rnorm(50, sd=15)))
```

## ANOVA table practice

The commonly accepted explanation for the extinction of the dinosaurs is a colossal asteroid strike. To support this hypothesis, we can look for iridium, a rare metal which is more common in asteroids, in rock samples. A sudden increase in iridium at some point in time supports the asteroid hypothesis.

We have data on 28 rock samples, recording the concentration of iridium (in ppb) for each sample, and the type of rock (Limestone or Shale). We use the model

$$Iridium = \beta_0 + \beta_1 IsShale + \varepsilon$$

We can use an ANOVA table to explore differences between these types of rock. Here is a partial ANOVA table for the model described above:

| Source | df | SS | MS | F |
| ---    | --- | ---|---|---|
| Model  |     |  88363  |   |   |
| Residual |   |    |   |   |
| Total  |     | 1252331   |   |   |

11. Fill in the remainder of the ANOVA table.


## Example data questions

How does stress impact animals' oxygen consumption? To investigate this question, biologists exposed 100 crabs to either 7.5 minutes of loud ship noise, or 7.5 minutes of normal ocean noise, and recorded their oxygen consumption. The data contain the following variables.

* `Mass`: crab mass (g)
* `Oxygen`: rate of oxygen consumption (micromoles per hour)
* `Noise`: type of noise (either `ocean` or `ship`)

The researchers are interested in the relationship between mass and oxygen consumption, and the impact of noise on this relationship. They make the following scatterplot:

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
crabs %>%
  ggplot(aes(x = Mass, y = Oxygen, 
             color = Noise, shape = Noise)) +
  geom_point(size=2) +
  theme_bw() +
  theme(text = element_text(size=20))
```

They consider three different models for this relationship ($\text{IsShip}$ is an indicator variable which is 1 for ship noise, and 0 for ocean noise):

$(1) \hspace{1cm} \text{oxygen} = \beta_0 + \beta_1 \text{mass} + \varepsilon$

$(2) \hspace{1cm} \text{oxygen} = \beta_0 + \beta_1 \text{IsShip} + \beta_2 \text{mass} + \varepsilon$

$(3) \hspace{1cm} \text{oxygen} = \beta_0 + \beta_1 \text{IsShip} + \beta_2 \text{mass} + \beta_3 \text{IsShip} \cdot \text{mass} +  \varepsilon$

The estimated equations for these models are

$(1) \hspace{1cm} \widehat{\text{oxygen}} = 161.29 + 0.23 \ \text{mass}$

$(2) \hspace{1cm} \widehat{\text{oxygen}} = 143.72 + 53.68 \ \text{IsShip} + 0.05 \ \text{mass}$

$(3) \hspace{1cm} \widehat{\text{oxygen}} = 52.06 + 257.38 \ \text{IsShip} + 1.92 \ \text{mass} - 4.05 \ \text{IsShip} \cdot \text{mass}$


12. Based on the scatterplot, which of these three models do you think is most appropriate? Explain your reasoning.

13. For each of the *other* two models, draw a scatterplot that would encourage you to choose that model, instead of the model selected in Question 12.

14. Using the estimated equation for the model you chose in question 12, what is the predicted rate of oxygen consumption for a 40g crab exposed to ocean noise? What about for a 40g crab exposed to ship noise?

15. What is the estimated change in oxygen consumption associated with a 1g increase in Mass, for crabs exposed to ship noise?

16. The researchers want to test whether there is any relationship between ship noise and oxygen consumption, after accounting for mass. Write down a null and alternative hypothesis for this test, in terms of one or more parameters from the model.

17. What test can the researchers use to test their hypotheses? Explain your reasoning, and provide the degrees of freedom.

18. Below are the ANOVA tables for the three fitted models. Use these tables to calculate a test statistic for the hypotheses in Question 16.

```{r}
m1 <- lm(Oxygen ~ Mass, data = crabs)
anova(m1)
```

<br/>

<br/>

```{r}
m2 <- lm(Oxygen ~ Mass + Noise, data = crabs)
anova(m2)
```

<br/>

<br/>

```{r}
m3 <- lm(Oxygen ~ Mass*Noise, data = crabs)
anova(m3)
```

19. Below is a picture of the distribution used to calculate a p-value with the test statistic in the previous question. Is the observed test statistic *usual* or *unusual* under the null hypothesis?

```{r, echo=F, message=F, warning=F}
x <- qf(ppoints(1000), df1 = 2, df2 = 96)
y <- df(x, df1 = 2, df2 = 96)
data.frame(x, y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  labs(x = "", y = "")
```


## More data practice

A Saint Olaf College student, undertook a study of the archery scores of students at the college who were enrolled in an archery course. Students taking the course record a score for each day they attend class from the first until the last day. We have the following data for each student in the class:

* `Attendance`: Number of days in class
* `Day1`: Archery score on first day
* `LastDay`: Archery score on last day

The student fits the following regression model:

```{r, eval=F}
m_archery <- lm(LastDay ~ Day1 + Attendance, data = ArcheryData)
summary(m_archery)
```

```
            Estimate Std. Error 
(Intercept) 113.1278   174.6178  
Day1          0.7578     0.2046  
Attendance   -1.3973     7.9871 

Multiple R-squared:  0.5127,	Adjusted R-squared:  0.4477 
```

20. Interpret each of the coefficients in the fitted model, in context of the data.

21. Interpret the $R^2$ for the model.

22. Calculate and interpret a 97% confidence interval for the change in archery score on the last day associated with a 1-point increase in archery score on the first day, holding attendance fixed. The critical value is $t^* = 2.40$.





